### 2013-04-30 paper.tex
Initial import.

\documentclass[prodmode]{acmlarge}

% Metadata Information
\acmVolume{0}
\acmNumber{0}
\acmArticle{0}
\articleSeq{0}
\acmYear{2013}
\acmMonth{5}

% Package to generate and customize Algorithm as per ACM style
\usepackage[ruled]{algorithm2e}
\SetAlFnt{\algofont}
\SetAlCapFnt{\algofont}
\SetAlCapNameFnt{\algofont}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}
\renewcommand{\algorithmcfname}{ALGORITHM}

% Page heads
\markboth{J.~Overbey}{Immutable Source-Mapped Abstract Syntax Tree}

% Title portion
\title{Immutable Source-Mapped Abstract Syntax Tree: \\
A Design Pattern for Refactoring Engine APIs}
\author{JEFFREY L.\ OVERBEY \affil{Auburn University}}

\begin{abstract}
TODO
\end{abstract}

\category{X.0.0}{TODO}{TODO}[TODO]

\terms{TODO}
\keywords{TODO}

\acmformat{Overbey, J. 2013. Immutable Source-Mapped Abstract Syntax Tree.}

\copyr{PLoP'13, October 23--26, Monticello, Illinois, USA. Copyright 2013 is
held by the author(s). ACM 978-1-4503-0107-7}

\begin{document}

\begin{bottomstuff}
This material is based upon work supported by the National Science Foundation
under Grant No.~CCF-1217271.

Permission to make digital or hard copies of all or part of this work for
personal or classroom use is granted without fee provided that copies are not
made or distributed for profit or commercial advantage and that copies bear
this notice and the full citation on the first page. To copy otherwise, to
republish, to post on servers or to redistribute to lists, requires prior
specific permission. A preliminary version of this paper was presented in a
writers' workshop at the 20th Conference on Pattern Languages of Programs
(PLoP).
\end{bottomstuff}

\maketitle

%\section{Abstract}
%\section{Example}
%\section{Context}
%\section{Problem}
%\section{Solution}
%\section{Structure}
%\section{Dynamics}
%\section{Implementation}
%\section{Example Resolved}
%\section{Variants}
%\section{Known Uses}
%\section{Consequences}
%\section{See Also}
%\section{Credits}

\section{Context}

\cite{TODO}

\section{Problem}

\section{Forces}

\section{Solution}
=> Rewrite text directly
=> Rewrite to new AST

\section{Example}

\section{Known Uses}
JDT
Red-Green Trees - MS
CDT
Clang
NetBeans???
gorf???

\section{Related Patterns and Guidelines}

\bibliographystyle{acmlarge}
\bibliography{paper}

% History dates
%\received{February 2009}{July 2009}{October 2009}

\end{document}

### 2013-05-19 paper.tex
First complete draft

@@
 \documentclass[prodmode]{acmlarge}
+\usepackage{graphicx}
+\usepackage{wrapfig}
+\usepackage{url}
+\newcommand{\ttt}[1]{\texttt{#1}}
+\newcommand{\code}[1]{\texttt{#1}}
 
@@
 \begin{abstract}
-TODO
+Many automated refactoring tools provide an extensibility mechanism, which
+allows third parties to contribute new refactorings.  However, implementing
+nontrivial refactorings requires deep syntactic knowledge about the program
+being refactored.  Refactoring tools almost always use abstract syntax trees
+(ASTs) to describe the structure of source code.  Therefore, it may be
+desirable to expose these trees via an API.  This pattern describes the common
+attributes that ASTs typically exhibit in order to be useful for implementing
+nontrivial refactorings while, at the same time, maintaining the
+characteristics of good API design.  This pattern is used in the Eclipse Java
+Development Tools' refactoring API as well as the Microsoft ``Roslyn'' CTP,
+among other uses.
 \end{abstract}
 
-\category{X.0.0}{TODO}{TODO}[TODO]
+\category{D.2.3}{Software Engineering}{Coding Tools and Techniques}
 
-\terms{TODO}
-\keywords{TODO}
+\terms{Languages}
+\keywords{abstract syntax trees, application programming interfaces, APIs,
+ASTs, immutability, refactoring}
 
@@
 
+Many integrated development environments (IDEs) provide extensibility
+mechanisms, which allow third-party developers to add new functionality to the
+IDE.  For IDEs that provide automated refactoring support, it is common to
+provide the ability to contribute new automated refactorings.
+
+However, automated refactorings manipulate source code, so implementing an
+automated refactoring requires a data structure that describes the structure of
+soruce code.  Refactoring tools almost always use a data structure called an
+\textit{abstract syntax tree} (AST).  ASTs describe source code hierarchically:
+e.g., classes contain methods, methods contain statements, statements contain
+expressions, etc.  ASTs are used in compilers and static analysis tools, but
+refactoring tools impose more stringent requirements on ASTs---specifically,
+the ASTs must be suitable for manipulating the user's source code.
+
+If an IDE provides an application programming interface (API) that allows third
+parties to contribute new refactorings, then it is often helpful for that API
+to provide access to ASTs as well.  This pattern considers ASTs exposed via
+refactoring engine extensibility APIs, identifying the common elements
+necessary to provide both a clean API and a data structure suitable for
+implementing nontrivial refactorings.
+
 \section{Context}
 
-\cite{TODO}
+You are implementing a refactoring engine or a similar tool that makes
+semi-automated edits to source code.  You have provided a mechanism that allows
+third parties to contribute new refactorings/transformations.  You have an API
+for adding, deleting, and changing source code files (so the mechanism to, say,
+``delete the first 27 characters of \textit{main.c}'' is well-defined).
 
@@
 
+Although you have an API for contributing refactorings/transformations, and you
+have provided the ability to make string-like edits to source code, this is a
+very primitive API that eschews any real knowledge about the syntactic
+structure of the language being refactored.
+
+Consider a simple Java refactoring that adds braces to the bodies of
+\textit{if} statements, if they do not already exist.  For example, it would
+change \\
+\centerline{\ttt{if (a < b) if (c < d) System.out.println("a");
+                            else System.out.println("b");}}
+into \\
+\centerline{\ttt{if (a < b) \{ if (c < d) \{ System.out.println("a"); \}
+                               else \{ System.out.println("b"); \} \}}.}
+Even this simple refactoring requires nontrivial knowledge of Java's syntax,
+including how the dangling \textit{else} is matched with a corresponding
+\textit{if}.
+
+How can the refactoring engine provide an API that exposes the syntactic
+structure of programs in a way that is useful for implementing refactorings?
+
 \section{Forces}
 
+\begin{itemize}
+
+\item The API must expose fine-grained details about the syntax of a program.
+For example, the preceding example requires being able to determine the
+presence or absence of braces \ttt{\{} \ttt{\}}, even though they are
+semantically irrelevant.
+
+\item Many refactorings are initiated only after the user selects a range of
+source code in a text editor.  Therefore, the API must allow clients to
+determine what syntactic construct(s) are selected.
+
+\item Clients must be able to use the API to manipulate source code in a way
+that is minimally invasive: refactorings that only affect a small, isolated
+region of the source code should not change any source code outside that
+region.  This includes preserving the program's comments and formatting.
+
+\item The API must follow general principles of good API design, including
+being easy to learn and difficult to misuse.
+
+\item The API should be usable when building refactorings for large code bases.
+Automated refactorings are most useful for maintaining such codes.
+
+\end{itemize}
+
 \section{Solution}
-=> Rewrite text directly
-=> Rewrite to new AST
 
-\section{Example}
+\textit{Use a parser for the language being refactored to construct an
+\textup{abstract syntax tree}, or \textup{AST}.  Make the AST
+\textup{immutable}---i.e., its structure cannot be changed---but provide
+\textup{source mapping} information so that every node in the AST can be mapped
+to a particular substring of the original source code.}
+
+\subsection{Abstract Syntax Trees (ASTs)}
+\label{ss:asts}
+
+(Readers familiar with ASTs may want to skip to \S\ref{ss:reqs}.)
+
+Programs have a natural hierarchical structure.  For example, a Java source
+file contains a package declaration, import statements, and one or more type
+declarations (i.e., class/interface declarations).  A type declaration contains
+field and method declarations.  A method's body consists of a sequence of
+statements.  And so forth.
+
+An \textit{abstract syntax tree} (AST) is a data structure that describes a
+particular program in terms of this hierarchy.  Continuing with the Java
+example, an AST typically describes the structure of a single Java source file.
+The root node has three child nodes, corresponding to the three parts of a Java
+source file: one child node describes the \ttt{package} statement, another
+contains a list of \ttt{import} statements, and a third contains a list of type
+declarations.  A type declaration has several child nodes, which describe
+(among other things) its visibility, the name of the class or interface, and
+the field/method declarations in its body.  A field declaration has child nodes
+describing its visibility, type, and so forth.
+
+\begin{wrapfigure}{R}{3in} % r is right, R is float right
+% Original image is 4.624 x 10.651 in
+\includegraphics[scale=0.6]{jdt-ast.eps}
+%\caption{Visualization of the Eclipse Java Development Tools' abstract syntax
+%tree for the following program: \\
+%\ \\
+%\centerline{\parbox{2in}{\ttt{\noindent public class Example \{ \\
+%\hspace*{.2em} public int FIVE = 5; \\
+%\}
+%}}} \\
+%\ \\
+%The highlighted node---a \textit{VariableDeclarationFragment}---corresponds to
+%the eight-character substring \ttt{FIVE = 5}, which begins with the 36th
+%character in the source code.
+%}
+\caption{Visualization of an AST from Eclipse JDT.}
+\label{fig:jdt-ast}
+\end{wrapfigure}
+
+Figure~\ref{fig:jdt-ast} shows a visualization of an AST for the following
+program:
+\vspace*{1em}
+
+\noindent \ttt{public class Example \{ \\
+\hspace*{.2em} public int FIVE = 5; \\
+\}
+}
+
+\vspace*{1em}
+\noindent (This visualization is produced by the Eclipse Java Development Tools
+(JDT), which are well known for their refactoring capabilities.)
+
+Notice how deeper levels of the tree describe progressively more fine-grained
+syntactic structures.  The root node represents the entire Java source file.
+The highlighted node---a \textit{VariableDeclarationFragment}---represents
+\ttt{FIVE = 5}, while its child nodes represent its constituent parts: the
+identifier \ttt{FIVE} and the numeric literal \ttt{5}.  These are leaf nodes,
+since these tokens (like keywords, identifiers, and numeric literals) represent
+the smallest syntactic units in the language.
+
+Developers familiar with the HTML or XML DOM (Document Object Model) will find
+this structure familiar.  The DOM is effectively an abstract syntax tree for
+HTML/XML documents.
+
+Deeper discussions on ASTs can be found in many introductory textbooks on
+compiler construction (e.g., \cite{aho06compilers} and
+\cite{torczon11engineering}).  Such books also discuss how parsers are
+constructed, including how they can be modified to construct ASTs.
+
+\subsection{Requirements of Refactoring ASTs}
+\label{ss:reqs}
+
+%The use of ASTs in refactoring tools is nearly universal.  
+%Since refactoring tools analyze and manipulate source code, the primary program
+%representation in a refactoring tool is always one that is fairly close to
+%source code---generally an abstract syntax tree.  The AST plays an absolutely
+%critical role.  It is used for both analysis and transformation, and every
+%refactoring makes use of it.
+
+ASTs are not unique to refactoring tools.  They were used in compilers long
+before refactoring tools existed.  They are also used in static analysis tools,
+prettyprinters, and other tools that analyze source code.  This makes reuse
+very tempting: If there is already a compiler for a language, and it contains
+an AST, why not reuse it in a refactoring tool?
+
+Unfortunately, an AST designed for use in another context (e.g., a compiler) is
+not necessarily suitable for refactoring.  This is because refactoring tools
+impose two very significant---and unusual---requirements on their ASTs.
+
+\begin{enumerate}
+\item To be useful for refactoring, an AST must accurately and
+precisely model the user's source code at a very fine level of detail.
+\item It must be possible to map nodes in the AST to locations in the
+source code, and vice versa.
+\end{enumerate}
+
+To illustrate the first requirement, consider an example from Fortran.  The
+Fortran language contains a number of input/output statements.  Two of these
+are \textsc{print} and \textsc{write}: the statements \ttt{print *, "X"} and
+\ttt{write (*,*) "X"} can be used interchangeably to print the string \ttt{X}
+to standard output.  In general, the statement \ttt{print} \textit{fmt}\ttt{,}
+\textit{expr} is equivalent to the statement \ttt{write (*,
+}\textit{fmt}\ttt{)} \textit{expr}.  Most compilers type check and translate
+them identically.  So, a compiler may choose to represent \textsc{print}
+statements as \textsc{write} statements in its AST.  The GNU Fortran compiler
+does this, for example.  However, the ability to convert \textsc{print}
+statements to \textsc{write} statements actually makes for a useful
+refactoring.  (Consider a program that uses \textsc{print} statements to write
+to standard output, but the programmer later needs to use \textsc{write}
+statements to write to a file instead.) To implement this refactoring, a
+refactoring tool needs to be able to distinguish \textsc{print} statements from
+\textsc{write} statements---a difficult task if they are represented
+identically in the AST.
+
+To explain the second requirement---that it must be possible to map nodes in
+the AST to locations in the source code, and vice versa---recall that many
+refactorings are initiated only after the user selects a range of source code
+in a text editor.  Usually, this selection is represented by two numbers: the
+\textit{offset} of the starting character and the \textit{length} of the
+selection.  In the simple Java program whose AST is shown in
+Figure~\ref{fig:jdt-ast}, the substring \ttt{FIVE = 5} is eight characters long
+and begins with the 36th character in the source code (offset 35, assuming the
+first character is considered to be at offset 0).  If the user selects the text
+\ttt{FIVE = 5} in the text editor, it must be easy to determine that this
+corresponds to the \textit{VariableDeclarationFragment} node in the AST.
+
+The easiest way to map AST nodes to source locations is to store source
+location information in the AST nodes themselves.  Again, consider the JDT AST
+shown in Figure~\ref{fig:jdt-ast}, and note the numbers in square brackets.
+The highlighted node is displayed as ``VariableDeclarationFragment [35, 8]'':
+this AST node corresponds to the eight-character string beginning at offset 35
+(\ttt{FIVE = 5}).  Its child node corresponding to the identifier \ttt{FIVE} is
+attributed with the offset-length pair [35, 4], while the numeric literal
+\ttt{5}'s source location is [42, 1].
+
+Now, consider what happens when a user selects text in the text editor.  This
+can be represented as an offset-length pair.  When source location information
+is available for every AST node, it becomes fairly easy to determine what
+syntactic construct corresponds to the text selection corresponds to: simply
+walk the tree, searching for a node whose source location aligns with the
+user's selection.
+
+Since AST nodes at deeper levels of the tree represent finer-grained syntactic
+constructs than their ancestor nodes, ASTs with source location information
+typically exhibit a nice property: the source location of a child node will
+always be contained within the source location of its parent.  This property
+can be used to expedite searches based on source location information.  For
+example, if a particular node corresponds to the source location [35, 8], and
+you are searching offset 63, then the desired node cannot possibly be a
+descendent of that node.
+
+%Preprocessors (such as the C preprocessor and M4) do not pose much of a problem
+%for compilers.  In a compiler, the preprocessor can be run first, and then its
+%output can serve as the input to the lexer/parser.  The preprocessor can be
+%completely unaware of what lexer/parser it is feeding into, and the
+%lexer/parser can be unaware of the fact that the code was preprocessed
+%(although it might handle directives like \code{\#line} to give better error
+%messages).
+%
+%Unfortunately, the C preprocessor makes refactoring very difficult; it is the
+%subject of Garrido's entire Ph.D.~thesis~\cite{garrido05program}.  The
+%compiler's approach of preprocesing the code first and then parsing it does not
+%work.  Since refactoring tools must modify the \textit{user's} source code,
+%they must be able to parse, analyze, and transform code \textit{with embedded
+%preprocessor directives.}
+%
+%This dissertation will not treat the topic of refactoring preprocessed code in
+%detail, but suffice it to say that a compiler's AST will almost certainly be
+%inadequate.  At a minimum, a refactoring must be able to determine which AST
+%nodes correspond to actual text in the user's source code and which ones are
+%the result of file inclusions and macro expansions.  For more details, see
+%Overbey et al.~\cite{wrt2009} and Garrido~\cite{garrido05program}.
+
+%\subsection{Mapping AST Nodes to Source Locations}
+%\label{ss:mapping}
+
+\subsection{Mutable vs.\ Immutable ASTs}
+\label{ss:manip}
+
+When ASTs are used in compilers, it is common to make them \textit{mutable},
+meaning that nodes can easily be added, changed, or deleted from the tree.  For
+example, a compiler might determine that \ttt{a = b*1} can be simplified to
+\ttt{a = b}, and then change the AST accordingly.
+
+At first, this seems like a reasonable mechanism for refactoring tools to use
+as well.  To remove a field declaration from a program, simply delete its
+\textit{FieldDeclaration} node from the AST (along with all its children).  To
+change the name of \ttt{FIVE}, change the contents of its
+\textit{SimpleNameNode} in the AST.  Of course, some mechanism is needed to
+prettyprint the modified AST, so these changes can be translated back into
+source code, but that is relatively straightforward to implement.
+
+Unfortunately, while a mutable AST can provide an elegant means of manipulating
+source code, that are other issues that can make mutable ASTs problematic from
+an API design perspective.  Some notable issues include the following.
+\begin{itemize}
+\item \textit{AST nodes contain source location information.}  After the
+structure of the AST is changed, what does the source location information in
+each node represent: the ``old'' source location (in the original code) or the
+``new'' location (after it has been modified)?  If it represents the old
+location, what should the source location be when new nodes are inserted into
+the tree?  If it represents the new location, then every time a change is made
+to the tree, source location information must be updated for every node in the
+tree, which may be expensive.
+\item \textit{AST nodes may be removed from the tree, leaving them orphaned.}
+Suppose a client traverses the AST, storing pointers to some ``interesting''
+nodes in a collection.  Then, an ancestor of one of those nodes is removed from
+the AST.  This will leave the client in a difficult-to-debug situation where
+his collection contains pointers to nodes that no longer exist.
+\item \textit{Semantic information may be invalidated.}  Refactorings often
+need to determine, for a particular use of a variable, what declaration that
+use corresponds to.  For example, in Java, if a field and a local variable were
+both named \ttt{x}, then in the expression \ttt{this.x = x}, the first \ttt{x}
+would refer to the field, while the second would refer to the local variable.
+However, if the local variable declaration were removed, the second \ttt{x}
+would refer to the field instead.  This can be confusing to clients, since
+information collected about the program at one point in time may no longer be
+true after the tree's structure changes.
+\item \textit{It may be possible to mutate the AST so that it does not
+represent a legal program.}  It might even be possible to mutate the AST so
+that it is no longer a tree at all.  In Java, classes can contain other
+classes.  What happens if a client tries to make a class an inner class of
+itself?  (This could turn the tree into a cyclic graph.)  The API for modifying
+the AST's structure could try to prevent such things, but attempting to detect
+and prevent every possible mistake would be difficult and potentially
+expensive.
+\item \textit{Mutability can be problematic in multithreaded code.}
+\end{itemize}
+
+Mutable ASTs were used in the original Smalltalk Refactoring
+Browser~\cite{roberts97refactoring}, as well as in
+CRefactory~\cite{garrido05program} and
+Photran (details of Photran's implementation are discussed
+in~\cite{sle2008}).  Many compilers use mutable ASTs internally.  However, these
+tools have one thing in common: the mutable AST is not accessible via any
+public API.  It is intended for internal use only, by developers who should be
+better equipped to handle all of the caveats.
+
+Guidelines for API design are discussed in detail in Jaroslav Tulach's
+\textit{Practical API Design}~\cite{tulach08practical}.  A presentation by
+Joshua Bloch~\cite{bloch} offers excellent advice as well.  Both authors
+suggest minimizing the amount of mutability in an API.  After all, mutable
+objects tend to have simpler interfaces, and they are thread-safe.  Both
+authors also emphasize that APIs should be easy to learn (even for programmers
+who do not read all of the API documentation) and difficult for clients to
+misuse.  Given these guidelines, and the list of liabilities for mutable ASTs
+given above, it seems preferable from the perspective of someone attempting to
+publish an API to avoid mutable ASTs.  But this raises a different question: If
+ASTs are immutable, what API should be provided for clients to manipulate
+source code?
+
+\subsection{Using Immutable ASTs for Source Code Manipulation}
+\label{ss:manip}
+
+The ultimate goal of a refactoring tool is to modify the user's source code.
+When ASTs are immutable, this cannot be done by modifying the structure of the
+AST.  Instead, it is done differently: \textit{Treat the source code as a
+string, and use the source location information in the AST to add, replace, and
+delete substrings.}
+
+As an example, consider deleting a field declaration.  In a mutable AST, you
+would simply remove the \textit{FieldDeclaration} node from the AST.  With an
+immutable AST, you would retrieve the source location mapping from the
+\textit{FieldDeclaration} node, and then remove that substring from the source
+file.
+
+In implementation, the code for removing a field declaration using a mutable
+AST would probably be something like \textit{field.removeFromTree()}.  The code
+for doing the same with an immutable AST would be something like
+\textit{sourceCode.deleteSubstring(field.getOffset(), field.getLength())}.
+
+Note that the manipulation code using a mutable AST is simpler, and it more
+directly expresses the programmer's intention.  Manipulating source code using
+offset/length information has a major disadvantage: it leads to ugly client
+code.  The client code can easily become littered with variables holding
+character offsets and methods that perform intricate string manipulations.
+
+In their discussions of API design, Bloch submits that it should be ``easy to
+read and maintain code that uses [the API]''~\cite{bloch}, and Tulach notes the
+advantages of APIs where the clients ``don't describe step-by-step what they
+want your API to do.  Rather, they `declare' what they want to have happen and
+then rely on your API to do it\dots''~[p.~225]\cite{tulach08practical}.
+
+So, is it possible to provide a cleaner source manipulation API, while still
+using an immutable AST?  It is, but some additional API is needed.
+
+One option is to provide an API that allows clients to specify what changes
+they want to make in terms of AST nodes.  This API can then translate these
+into string edits.  For example, an API might offer a \textit{Rewriter} class.
+The client would inform the \textit{Rewriter} that it wants to effectively
+delete a particular \textit{FieldDeclaration} node from the AST:
+\textit{rewriter.delete(field)}.  When it is ready to manipulate the source
+code, it delegates this task to the rewriter:
+\textit{rewriter.rewrite(sourceCode)}.
+
+A second option is to work with ASTs in a functional style, where every edit
+returns a new AST.  So, a call to \textit{field.removeFromTree()} would not
+change the existing AST, but rather it would return a new AST that has the
+given field omitted.  (Of course, this requires a very careful implementation.
+Memory consumption and copying are a concern, particularly when ASTs are large.
+Also, a mechanism must be provided so that clients can easily map nodes in the
+original tree to the ``same'' node in the modified tree, and vice versa.)
+
+%\section{Example}
 
 \section{Known Uses}
-JDT
-Red-Green Trees - MS
-CDT
-Clang
-NetBeans???
-gorf???
 
-\section{Related Patterns and Guidelines}
+Immutable, source-mapped ASTs are used by the Eclipse Java Development Tools
+(JDT) and C/C++ Development Tools (CDT), the Microsoft ``Roslyn'' Community
+Technology Preview (CTP), Clang, and the original refactoring engine in Apple
+Xcode.  However, while all use immutable ASTs with source mapping information,
+the actual rewriting APIs vary substantially.
+
+Both Clang and the original refactoring engine in Xcode~3.0 modify source code
+using offset-length information directly.  Notably, Clang offers a
+\textit{RewriteRope} class---essentially, a custom string class that allows for
+efficient mid-string insertions and deletions~\cite{clang-rewriterope}.
+(It uses a B-tree structure inspired by, but not identical to, that described
+in the original paper on Ropes~\cite{boehm95ropes}.)
+
+The Eclipse Platform is designed to be language-independent; individual
+plug-ins like JDT and CDT determine what languages it supports.  The Eclipse
+Platform includes a component called the Eclipse Language Toolkit
+(LTK)~\cite{ltk}.  which provides (among other things) a language-independent
+API for manipulating text files using offset/length information.  Specifically,
+\textit{TextEdit} objects describe individual changes (like ``delete three
+characters starting at offset 462''), and many such changes are aggregated into
+a single \textit{Change} object that represents the net effect of a
+refactoring.  Of course, the LTK provides the infrastructure necessary to apply
+\textit{Change} objects, writing the modifications to disk, as well as to undo
+those changes at a later point in time.
+
+Eclipse JDT provides a class (\textit{ASTRewrite}) that allows the programmer
+to specify modifications in terms of Java AST nodes, and then translates those
+changes into an LTK \textit{TextEdit} object~\cite{jdt-astrewrite}.  Eclipse
+CDT's refactoring engine is based on JDT's and follows a similar design; it
+offers its own version of \textit{ASTRewrite}~\cite{cdt-refactoring}.
+
+Microsoft's ``Roslyn'' CTP takes a functional approach, where imperative
+methods on AST nodes do not modify the AST but rather return a new
+AST~\cite{vogel12roslyn}.  Since a na\"{i}ve implementation could make this
+prohibitively expensive, the Roslyn team devised an implementation that
+they call \textit{red-green trees} (named after the colors of the
+whiteboard markers that were used during the design
+meeting)~\cite{lippert12persistence}.  Essentially, the AST exposed via the API
+(the ``red tree'') is a fa\,{c}ade for a different, internal tree (the ``green
+tree''); the implementation attempts to reuse existing AST nodes whenever
+possible (which is possible since all nodes are immutable).
+
+% TODO: Preprocessor complications
+% TODO: Example code?
+
+%\section{Related Patterns and Guidelines}
 

### 2013-07-14 paper.tex
Second draft

@@
-\documentclass[prodmode]{acmlarge}
+% TODO: Version with [draft] is for draft mode only
+%\documentclass[prodmode]{acmlarge}
+\documentclass[draft]{acmlarge}
 \usepackage{graphicx}
@@
 \usepackage{url}
+\usepackage{datetime} % for \today
+\usepackage{listings} % for lstlisting environment
+\usepackage{placeins} % for \FloatBarrier
 \newcommand{\ttt}[1]{\texttt{#1}}
@@
 \title{Immutable Source-Mapped Abstract Syntax Tree: \\
-A Design Pattern for Refactoring Engine APIs}
+A Design Pattern for Refactoring Engine APIs \\ \today}
 \author{JEFFREY L.\ OVERBEY \affil{Auburn University}}
@@
 automated refactoring requires a data structure that describes the structure of
-soruce code.  Refactoring tools almost always use a data structure called an
+source code.  Refactoring tools almost always use a data structure called an
 \textit{abstract syntax tree} (AST).  ASTs describe source code hierarchically:
@@
 expressions, etc.  ASTs are used in compilers and static analysis tools, but
-refactoring tools impose more stringent requirements on ASTs---specifically,
-the ASTs must be suitable for manipulating the user's source code.
+refactoring tools impose more stringent requirements---specifically, the ASTs
+must be suitable for manipulating the user's source code.
 
@@
 You are implementing a refactoring engine or a similar tool that makes
-semi-automated edits to source code.  You have provided a mechanism that allows
+automated edits to source code.  You have provided a mechanism that allows
 third parties to contribute new refactorings/transformations.  You have an API
@@
 \section{Problem}
+\label{sec:Problem}
 
@@
 \textit{if} statements, if they do not already exist.  For example, it would
-change \\
-\centerline{\ttt{if (a < b) if (c < d) System.out.println("a");
-                            else System.out.println("b");}}
-into \\
-\centerline{\ttt{if (a < b) \{ if (c < d) \{ System.out.println("a"); \}
-                               else \{ System.out.println("b"); \} \}}.}
+make the following change.
+
+\begin{center}
+\vspace*{1em}
+\begin{tabular}{ccc}
+\begin{minipage}{2.25in}
+\texttt{%
+                if (a < b) \\
+\hspace*{1.5em}   if (c < d) \\
+\hspace*{3.0em}     System.out.println("a"); \\
+\hspace*{1.5em}   else \\
+\hspace*{3.0em}     System.out.println("b");} \\
+ \\
+\end{minipage}
+& $\Longrightarrow$ &
+\begin{minipage}{2.25in}
+\texttt{%
+                if (a < b) \textbf{\{} \\
+\hspace*{1.5em}   if (c < d) \textbf{\{} \\
+\hspace*{3.0em}     System.out.println("a"); \\
+\hspace*{1.5em}   \textbf{\}} else \textbf{\{} \\
+\hspace*{3.0em}     System.out.println("b"); \\
+\hspace*{1.5em}   \textbf{\}} \\
+                \textbf{\}}}
+\end{minipage}
+\end{tabular}
+\vspace*{1em}
+\end{center}
+
 Even this simple refactoring requires nontrivial knowledge of Java's syntax,
@@
 
-\subsection{Abstract Syntax Trees (ASTs)}
+\subsection{Abstract Syntax Trees}
 \label{ss:asts}
+
+\begin{wrapfigure}{R}{3in} % r is right, R is float right
+% Original image is 4.624 x 10.651 in
+\includegraphics[scale=0.65]{jdt-ast.eps}
+%\caption{Visualization of the Eclipse Java Development Tools' abstract syntax
+%tree for the following program: \\
+%\ \\
+%\centerline{\parbox{2in}{\ttt{\noindent public class Example \{ \\
+%\hspace*{.2em} public int FIVE = 5; \\
+%\}
+%}}} \\
+%\ \\
+%The highlighted node---a \textit{VariableDeclarationFragment}---corresponds to
+%the eight-character substring \ttt{FIVE = 5}, which begins with the 36th
+%character in the source code.
+%}
+\caption{Visualization of an AST from Eclipse JDT.}
+\label{fig:jdt-ast}
+\end{wrapfigure}
 
@@
 describing its visibility, type, and so forth.
-
-\begin{wrapfigure}{R}{3in} % r is right, R is float right
-% Original image is 4.624 x 10.651 in
-\includegraphics[scale=0.6]{jdt-ast.eps}
-%\caption{Visualization of the Eclipse Java Development Tools' abstract syntax
-%tree for the following program: \\
-%\ \\
-%\centerline{\parbox{2in}{\ttt{\noindent public class Example \{ \\
-%\hspace*{.2em} public int FIVE = 5; \\
-%\}
-%}}} \\
-%\ \\
-%The highlighted node---a \textit{VariableDeclarationFragment}---corresponds to
-%the eight-character substring \ttt{FIVE = 5}, which begins with the 36th
-%character in the source code.
-%}
-\caption{Visualization of an AST from Eclipse JDT.}
-\label{fig:jdt-ast}
-\end{wrapfigure}
 
@@
 
-\subsection{Requirements of Refactoring ASTs}
+\subsection{Fulfilling the Requirements of Refactoring}
 \label{ss:reqs}
@@
 \begin{enumerate}
-\item To be useful for refactoring, an AST must accurately and
-precisely model the user's source code at a very fine level of detail.
+\item To be useful for refactoring, an AST must accurately and precisely model
+the user's source code at a very fine level of detail.  It may even be
+necessary to include information about punctuation and comments in order to
+refactor code appropriately~\cite{sommerlad08retaining}.
 \item It must be possible to map nodes in the AST to locations in the
@@
 meaning that nodes can easily be added, changed, or deleted from the tree.  For
-example, a compiler might determine that \ttt{a = b*1} can be simplified to
-\ttt{a = b}, and then change the AST accordingly.
+example, a compiler might determine that \ttt{a~=~b*1} can be simplified to
+\ttt{a~=~b}, and then change the AST accordingly.
 
@@
 to the tree, source location information must be updated for every node in the
-tree, which may be expensive.
+tree, which may be expensive \dots or else a clever implementation (e.g.,
+computation from relative offsets) must be used instead.
 \item \textit{AST nodes may be removed from the tree, leaving them orphaned.}
@@
 
-Note that the manipulation code using a mutable AST is simpler, and it more
+The situation gets a bit worse when changes are made to nested constructs.
+Recall the example from Section~\ref{sec:Problem}, which added braces to two
+\textit{if} statements, one nested inside the other.  Suppose the inner
+\textit{if} statement were modified first: Adding braces to it would change the
+position at which the closing brace for the outer \textit{if} statement would
+need to be added.  Likewise, if braces were added to the outer \textit{if}
+statement first, then this would shift the location of the inner \textit{if}
+statement.  In other words, when refactorings affect nested language
+constructs, offset/length computations can become complex.
+
+So, manipulating source code using a mutable AST is simpler, and it more
 directly expresses the programmer's intention.  Manipulating source code using
-offset/length information has a major disadvantage: it leads to ugly client
-code.  The client code can easily become littered with variables holding
-character offsets and methods that perform intricate string manipulations.
+offset/length information leads to ugly client code.  The client code can
+easily become littered with variables holding character offsets and methods
+that perform intricate string manipulations.
 

### 2013-08-04 paper.tex
Modified options, layout for resubmission

@@
 % TODO: Version with [draft] is for draft mode only
-%\documentclass[prodmode]{acmlarge}
-\documentclass[draft]{acmlarge}
+% Also comment/uncomment \today in subtitle when in draft mode
+\documentclass[prodmode]{acmlarge}
+%\documentclass[draft]{acmlarge}
 \usepackage{graphicx}
@@
 \title{Immutable Source-Mapped Abstract Syntax Tree: \\
-A Design Pattern for Refactoring Engine APIs \\ \today}
+A Design Pattern for Refactoring Engine APIs} % \\ \today}
 \author{JEFFREY L.\ OVERBEY \affil{Auburn University}}
@@
 
-\begin{wrapfigure}{R}{3in} % r is right, R is float right
+(Readers familiar with ASTs may want to skip to \S\ref{ss:reqs}.)
+
+Programs have a natural hierarchical structure.  For example, a Java source
+file contains a package declaration, import statements, and one or more type
+declarations (i.e., class/interface declarations).  A type declaration contains
+field and method declarations.  A method's body consists of a sequence of
+statements.  And so forth.
+
+An \textit{abstract syntax tree} (AST) is a data structure that describes a
+particular program in terms of this hierarchy.  Continuing with the Java
+example, an AST typically describes the structure of a single Java source file.
+The root node has three child nodes, corresponding to the three parts of a Java
+source file: one child node describes the
+
+\begin{wrapfigure}{r}{3in} % r is right, R is float right
 % Original image is 4.624 x 10.651 in
-\includegraphics[scale=0.65]{jdt-ast.eps}
+\includegraphics[scale=0.62]{jdt-ast.eps}
 %\caption{Visualization of the Eclipse Java Development Tools' abstract syntax
@@
 
-(Readers familiar with ASTs may want to skip to \S\ref{ss:reqs}.)
-
-Programs have a natural hierarchical structure.  For example, a Java source
-file contains a package declaration, import statements, and one or more type
-declarations (i.e., class/interface declarations).  A type declaration contains
-field and method declarations.  A method's body consists of a sequence of
-statements.  And so forth.
-
-An \textit{abstract syntax tree} (AST) is a data structure that describes a
-particular program in terms of this hierarchy.  Continuing with the Java
-example, an AST typically describes the structure of a single Java source file.
-The root node has three child nodes, corresponding to the three parts of a Java
-source file: one child node describes the \ttt{package} statement, another
+\noindent
+\ttt{package} statement, another
 contains a list of \ttt{import} statements, and a third contains a list of type
@@
 constructed, including how they can be modified to construct ASTs.
+
+\FloatBarrier
 

### 2013-10-01 paper.tex
Final revisions

@@
 \newcommand{\code}[1]{\texttt{#1}}
+\newcommand{\jb}[1]{\fbox{\texttt{#1}\rule[-1pt]{0pt}{.9em}}}
+\newcommand{\kb}[1]{\mbox{\texttt{\small #1}\rule{7.3pt}{0pt}\rule[-1pt]{0pt}{.9em}}}
+\newcommand{\lb}[1]{\mbox{\texttt{\small #1}\rule{3.2pt}{0pt}\rule[-1pt]{0pt}{.9em}}}
 
@@
 \begin{abstract}
-Many automated refactoring tools provide an extensibility mechanism, which
+Many automated refactoring tools provide an extensibility mechanisms that
 allows third parties to contribute new refactorings.  However, implementing
@@
 parties to contribute new refactorings, then it is often helpful for that API
-to provide access to ASTs as well.  This pattern considers ASTs exposed via
-refactoring engine extensibility APIs, identifying the common elements
-necessary to provide both a clean API and a data structure suitable for
-implementing nontrivial refactorings.
+to provide access to ASTs as well.  This pattern considers ASTs exposed by such
+APIs, identifying the common elements necessary to provide both a clean API and
+a data structure suitable for implementing nontrivial refactorings.
 
@@
 region of the source code should not change any source code outside that
-region.  This includes preserving the program's comments and formatting.
+region.  This includes preserving the program's comments and
+formatting~\cite{sommerlad08retaining}.
 
@@
 %}
-\caption{Visualization of an AST from Eclipse JDT.}
+\caption{Visualization of an AST produced by the Eclipse Java Development Tools
+(JDT).}
 \label{fig:jdt-ast}
@@
 \vspace*{1em}
-\noindent (This visualization is produced by the Eclipse Java Development Tools
-(JDT), which are well known for their refactoring capabilities.)
-
 Notice how deeper levels of the tree describe progressively more fine-grained
@@
 the AST.  This will leave the client in a difficult-to-debug situation where
-his collection contains pointers to nodes that no longer exist.
+his collection contains pointers to nodes that no longer exist in the AST.
 \item \textit{Semantic information may be invalidated.}  Refactorings often
@@
 Browser~\cite{roberts97refactoring}, as well as in
-CRefactory~\cite{garrido05program} and
-Photran (details of Photran's implementation are discussed
-in~\cite{sle2008}).  Many compilers use mutable ASTs internally.  However, these
-tools have one thing in common: the mutable AST is not accessible via any
-public API.  It is intended for internal use only, by developers who should be
-better equipped to handle all of the caveats.
+CRefactory~\cite{garrido05program} and Photran~\cite{sle2008}.  Many compilers
+use mutable ASTs internally.  However, these tools have one thing in common:
+the mutable AST is not accessible via any public API.  It is intended for
+internal use only, by developers who should be better equipped to handle all of
+the caveats.
 
@@
 Joshua Bloch~\cite{bloch} offers excellent advice as well.  Both authors
-suggest minimizing the amount of mutability in an API.  After all, mutable
+suggest minimizing the amount of mutability in an API.  After all, immutable
 objects tend to have simpler interfaces, and they are thread-safe.  Both
@@
 Recall the example from Section~\ref{sec:Problem}, which added braces to two
-\textit{if} statements, one nested inside the other.  Suppose the inner
+\textit{if} statements, one nested inside the other.  Suppose the outer
 \textit{if} statement were modified first: Adding braces to it would change the
-position at which the closing brace for the outer \textit{if} statement would
-need to be added.  Likewise, if braces were added to the outer \textit{if}
-statement first, then this would shift the location of the inner \textit{if}
-statement.  In other words, when refactorings affect nested language
-constructs, offset/length computations can become complex.
+position at which the closing brace for the inner \textit{if} statement would
+need to be added.  Consider a somewhat simplified example.  Pictorially:
+
+\begin{center}
+\jb{i}\jb{f}\jb{(}\jb{1}\jb{)}\jb{ }%
+\jb{i}\jb{f}\jb{(}\jb{2}\jb{)}\jb{ }%
+\jb{s}\jb{(}\jb{)}\jb{;}\\
+\lb{\,}\kb{0}\kb{1}\kb{2}\kb{3}\kb{4}\kb{5}\kb{6}\kb{7}\kb{8}\lb{9}%
+\lb{10}\lb{11}\lb{12}\lb{13}\lb{14}\lb{15} \\
+$\Downarrow$ \rule[-0.5em]{0pt}{1.2em} \\
+\jb{i}\jb{f}\jb{(}\jb{1}\jb{)}\jb{ }%
+\jb{\{}\jb{ }%
+\jb{i}\jb{f}\jb{(}\jb{2}\jb{)}\jb{ }%
+\jb{s}\jb{(}\jb{)}\jb{;}%
+\jb{ }\jb{\}}\\
+\lb{\,}\kb{0}\kb{1}\kb{2}\kb{3}\kb{4}\kb{5}\kb{6}\kb{7}\kb{8}\lb{9}%
+\lb{10}\lb{11}\lb{12}\lb{13}\lb{14}\lb{15}\lb{16}\lb{17}\lb{18}\lb{19} \\
+\end{center}
+
+\noindent
+The body of the inner \textit{if} statement begins at offset~12 in the original
+code.  But after adding braces to the outer \textit{if} statement, the body of
+the inner \textit{if} statement is shifted to offset~14.
+
+%Likewise, if braces were added to the inner \textit{if} statement first, then
+%this would shift the location of the closing brace for the outer \textit{if}
+%statement.
+In other words, when refactorings affect nested language constructs,
+offset/length computations can become complex, since the edits for one
+construct may interact with the edits for another construct.
 
@@
 code, it delegates this task to the rewriter:
-\textit{rewriter.rewrite(sourceCode)}.
+\textit{rewriter.rewrite(sourceCode)}.  This is illustrated in
+Figure~\ref{fig:seq-diagram}.  The Eclipse Java and C/C++ Development
+Tools~\cite{jdt-astrewrite,cdt-refactoring} take this approach
+(with slightly different class and method names).
+
+\begin{figure}[!bt]
+\begin{center}
+\includegraphics[width=\textwidth]{seq-diagram.eps}
+\end{center}
+\caption{Sequence diagram illustrating use of a Rewriter to modify source code.}
+\label{fig:seq-diagram}
+\end{figure}
 
@@
 Immutable, source-mapped ASTs are used by the Eclipse Java Development Tools
-(JDT) and C/C++ Development Tools (CDT), the Microsoft ``Roslyn'' Community
-Technology Preview (CTP), Clang, and the original refactoring engine in Apple
-Xcode.  However, while all use immutable ASTs with source mapping information,
-the actual rewriting APIs vary substantially.
+(JDT) and C/C++ Development Tools (CDT), as well as the Scala IDE for Eclipse,
+the Microsoft ``Roslyn'' Community Technology Preview (CTP), Clang, and the
+original refactoring engine in Apple Xcode.  However, while all use immutable
+ASTs with source mapping information, the actual rewriting APIs vary
+substantially.
 
@@
 
+Refactoring support in the Scala IDE for
+Eclipse~\cite{scalaide,scalarefactoring} takes a functional approach,
+implementing refactorings as tree transformations.  Source generation is
+handled separately: one printer uses position information to retain the
+formatting of the original code, while another pretty-prints new nodes that
+have been added to the AST~\cite{stocker10scala}.
+
 Microsoft's ``Roslyn'' CTP takes a functional approach, where imperative
@@
 
+\textit{The author would like to thank Peter Sommerlad for the invaluable
+feedback and suggestions he provided as the shepherd for this paper.}
+
 \bibliographystyle{acmlarge}

### 2013-10-01 paper.tex
Fixed references

@@
 Figure~\ref{fig:seq-diagram}.  The Eclipse Java and C/C++ Development
-Tools~\cite{jdt-astrewrite,cdt-refactoring} take this approach
-(with slightly different class and method names).
+Tools~\cite{cdt-refactoring} take this approach (with slightly different class
+and method names).
 
@@
 \begin{center}
-\includegraphics[width=\textwidth]{seq-diagram.eps}
+\includegraphics[width=.9\textwidth]{seq-diagram.eps}
 \end{center}
@@
 \textit{RewriteRope} class---essentially, a custom string class that allows for
-efficient mid-string insertions and deletions~\cite{clang-rewriterope}.
+efficient mid-string insertions and
+deletions\footnote{\url{http://clang.llvm.org/doxygen/classclang_1_1RewriteRope.html}}.
 (It uses a B-tree structure inspired by, but not identical to, that described
@@
 
-Eclipse JDT provides a class (\textit{ASTRewrite}) that allows the programmer
+Eclipse JDT provides a class (\textit{ASTRewrite}\footnote{\url{http://help.eclipse.org/kepler/topic/org.eclipse.jdt.doc.isv/reference/api/org/eclipse/jdt/core/dom/rewrite/ASTRewrite.html}}) that allows the programmer
 to specify modifications in terms of Java AST nodes, and then translates those
-changes into an LTK \textit{TextEdit} object~\cite{jdt-astrewrite}.  Eclipse
-CDT's refactoring engine is based on JDT's and follows a similar design; it
-offers its own version of \textit{ASTRewrite}~\cite{cdt-refactoring}.
+changes into an LTK \textit{TextEdit} object.  Eclipse CDT's refactoring engine
+is based on JDT's and follows a similar design; it offers its own version of
+textit{ASTRewrite}~\cite{cdt-refactoring}.
 
 Refactoring support in the Scala IDE for
-Eclipse~\cite{scalaide,scalarefactoring} takes a functional approach,
-implementing refactorings as tree transformations.  Source generation is
-handled separately: one printer uses position information to retain the
-formatting of the original code, while another pretty-prints new nodes that
-have been added to the AST~\cite{stocker10scala}.
+Eclipse\footnote{\url{http://scala-refactoring.org},
+\url{http://scala-ide.org/}} takes a functional approach, implementing
+refactorings as tree transformations.  Source generation is handled separately:
+one printer uses position information to retain the formatting of the original
+code, while another pretty-prints new nodes that were added to the
+AST~\cite{stocker10scala}.
 

### 2014-05-01 paper.tex
Camera-ready

@@
 % Title portion
-\title{Immutable Source-Mapped Abstract Syntax Tree: \\
+\title{Immutable Source-Mapped Abstract Syntax Tree:\ \\
 A Design Pattern for Refactoring Engine APIs} % \\ \today}
@@
 nontrivial refactorings while, at the same time, maintaining the
-characteristics of good API design.  This pattern is used in the Eclipse Java
-Development Tools' refactoring API as well as the Microsoft ``Roslyn'' CTP,
-among other uses.
+characteristics of good API design.  In particular, these trees tend to have an
+immutable structure, and they maintain very detailed information about the
+source code, including exact character positions.  This pattern is used in the
+Eclipse Java Development Tools' refactoring API as well as the Microsoft
+``Roslyn'' CTP, among other uses.
 \end{abstract}
@@
 \terms{Languages}
-\keywords{abstract syntax trees, application programming interfaces, APIs,
-ASTs, immutability, refactoring}
+\keywords{abstract syntax trees, APIs, ASTs, immutability, refactoring}
 
@@
 \begin{bottomstuff}
-This material is based upon work supported by the National Science Foundation
-under Grant No.~CCF-1217271.
-
 Permission to make digital or hard copies of all or part of this work for
@@
 writers' workshop at the 20th Conference on Pattern Languages of Programs
-(PLoP).
+(PLoP).  PLoP'13, October 23--26, Monticello, Illinois, USA. Copyright 2013 is
+held by the author(s). HILLSIDE 978-1-941652-00-8
 \end{bottomstuff}
@@
 \caption{Visualization of an AST produced by the Eclipse Java Development Tools
-(JDT).}
+(JDT).\vspace*{-4em}}
 \label{fig:jdt-ast}
@@
 program:
-\vspace*{1em}
+\vspace*{.5em}
 
@@
 
-\vspace*{1em}
+\vspace*{.5em}
 Notice how deeper levels of the tree describe progressively more fine-grained
@@
 Deeper discussions on ASTs can be found in many introductory textbooks on
-compiler construction (e.g., \cite{aho06compilers} and
-\cite{torczon11engineering}).  Such books also discuss how parsers are
+compiler construction, such as \cite{aho06compilers} and
+\cite{torczon11engineering}.  Such books also discuss how parsers are
 constructed, including how they can be modified to construct ASTs.
@@
 Joshua Bloch~\cite{bloch} offers excellent advice as well.  Both authors
-suggest minimizing the amount of mutability in an API.  After all, immutable
-objects tend to have simpler interfaces, and they are thread-safe.  Both
-authors also emphasize that APIs should be easy to learn (even for programmers
-who do not read all of the API documentation) and difficult for clients to
-misuse.  Given these guidelines, and the list of liabilities for mutable ASTs
-given above, it seems preferable from the perspective of someone attempting to
-publish an API to avoid mutable ASTs.  But this raises a different question: If
-ASTs are immutable, what API should be provided for clients to manipulate
-source code?
+suggest minimizing the amount of mutability in an API.  Immutable objects tend
+to have simpler interfaces, and they are thread-safe.  Both authors also
+emphasize that APIs should be easy to learn (even for programmers who do not
+read all of the API documentation) and difficult for clients to misuse.  Given
+these guidelines, and the list of liabilities for mutable ASTs given above, it
+seems preferable from the perspective of someone attempting to publish an API
+to avoid mutable ASTs.  But this raises a different question: If ASTs are
+immutable, what API should be provided for clients to manipulate source code?
 
@@
 want your API to do.  Rather, they `declare' what they want to have happen and
-then rely on your API to do it\dots''~[p.~225]\cite{tulach08practical}.
+then rely on your API to do it\dots''~\cite[p.~225]{tulach08practical}.
 
@@
 is based on JDT's and follows a similar design; it offers its own version of
-textit{ASTRewrite}~\cite{cdt-refactoring}.
+\textit{ASTRewrite}~\cite{cdt-refactoring}.
 
@@
 meeting)~\cite{lippert12persistence}.  Essentially, the AST exposed via the API
-(the ``red tree'') is a fa\,{c}ade for a different, internal tree (the ``green
+(the ``red tree'') is a fa\c{c}ade for a different, internal tree (the ``green
 tree''); the implementation attempts to reuse existing AST nodes whenever
@@
 
-\textit{The author would like to thank Peter Sommerlad for the invaluable
-feedback and suggestions he provided as the shepherd for this paper.}
+\section*{Acknowledgments}
+This material is based upon work supported by the National Science Foundation
+under Grant No.~CCF-1217271.  The author would like to thank Peter Sommerlad,
+who shepherded this pattern, as well as the members of Workshop~000001 at
+PLoP~2013: Filipe Correia, Dibyendu Goswami, Joshua Kerievsky, Juan Reza, and
+Russ Rubis.  Their feedback was immensely helpful.
 

### 2016-07-13 paper.tex
Fixed copyright footer
Prepare for TPLoP revisions
Updated Problems section to use reverse conditional.

@@
 A Design Pattern for Refactoring Engine APIs} % \\ \today}
-\author{JEFFREY L.\ OVERBEY \affil{Auburn University}}
+\author{JEFFREY L.\ OVERBEY and CARL R.\ WORLEY \affil{Auburn University}}
 
 \begin{abstract}
-Many automated refactoring tools provide an extensibility mechanisms that
-allows third parties to contribute new refactorings.  However, implementing
+Many automated refactoring tools provide extensibility mechanisms that
+allow third parties to contribute new refactorings.  However, implementing
 nontrivial refactorings requires deep syntactic knowledge about the program
@@
 
-\copyr{PLoP'13, October 23--26, Monticello, Illinois, USA. Copyright 2013 is
-held by the author(s). ACM 978-1-4503-0107-7}
+\copyr{}%PLoP'13, October 23--26, Monticello, Illinois, USA. Copyright 2013 is
+%held by the author(s). ACM 978-1-4503-0107-7}
 
@@
 
-Although you have an API for contributing refactorings/transformations, and you
-have provided the ability to make string-like edits to source code, this is a
-very primitive API that eschews any real knowledge about the syntactic
-structure of the language being refactored.
+Although you have an API for contributing refactorings/transformations, it is 
+very primitive, only capable of string-like manipulations. This API lacks any
+sophisticated knowledge of the text being edited.
 
-Consider a simple Java refactoring that adds braces to the bodies of
-\textit{if} statements, if they do not already exist.  For example, it would
-make the following change.
+Consider a simple Java refactoring that reverses the conditional meaning of
+\textit{if} statements. For example, it would make the following change.
 
@@
 \texttt{%
-                if (a < b) \\
-\hspace*{1.5em}   if (c < d) \\
-\hspace*{3.0em}     System.out.println("a"); \\
-\hspace*{1.5em}   else \\
-\hspace*{3.0em}     System.out.println("b");} \\
+                  if (!(a < b)) \textbf{\{}\\
+\hspace*{1.5em}     System.out.println("a"); \\
+                  \textbf{\}} else \textbf{\{}\\
+\hspace*{1.5em}     System.out.println("b"); \\
+                  \textbf{\}}} 
  \\
@@
                   if (a < b) \textbf{\{} \\
-\hspace*{1.5em}   if (c < d) \textbf{\{} \\
-\hspace*{3.0em}     System.out.println("a"); \\
-\hspace*{1.5em}   \textbf{\}} else \textbf{\{} \\
-\hspace*{3.0em}     System.out.println("b"); \\
-\hspace*{1.5em}   \textbf{\}} \\
-                \textbf{\}}}
+\hspace*{1.5em}     System.out.println("b"); \\
+                  \textbf{\}} else \textbf{\{} \\
+\hspace*{1.5em}     System.out.println("a"); \\
+                  \textbf{\}} \\
+                }
 \end{minipage}
@@
 
-Even this simple refactoring requires nontrivial knowledge of Java's syntax,
-including how the dangling \textit{else} is matched with a corresponding
-\textit{if}.
+To perform this refactoring, the conditional expression must be reversed in
+meaning and the contents of the \textit{if} and \textit{else} branches must be
+interchanged. Changing the condition requires removing the \textit{not} and the 
+redundant parentheses. The former is simple enough, but the latter requires 
+knowledge of matching parentheses. Interchanging the branches similarly requires 
+knowledge of the matching braces, the location of the \textit{if} and dangling 
+\textit{else}, and the presence or absence of any statements that alter flow of 
+control such as \textit{goto}. In essence, knowledge of the program's syntax is 
+required.
 
-How can the refactoring engine provide an API that exposes the syntactic
-structure of programs in a way that is useful for implementing refactorings?
+\textit{How can the refactoring engine provide an API that exposes the syntactic
+structure of programs in a way that is useful for implementing refactorings?}
 
@@
 For example, the preceding example requires being able to determine the
-presence or absence of braces \ttt{\{} \ttt{\}}, even though they are
-semantically irrelevant.
+exact location of parentheses \ttt{(} \ttt{)}.
 

### 2016-07-14 paper.tex
Variants are now discussed more explicity.

@@
 A Design Pattern for Refactoring Engine APIs} % \\ \today}
-\author{JEFFREY L.\ OVERBEY and CARL R.\ WORLEY \affil{Auburn University}}
+\author{JEFFREY L.\ OVERBEY \and CARL R.\ WORLEY \affil{Auburn University}}
 
@@
 
-When ASTs are used in compilers, it is common to make them \textit{mutable},
-meaning that nodes can easily be added, changed, or deleted from the tree.  For
-example, a compiler might determine that \ttt{a~=~b*1} can be simplified to
-\ttt{a~=~b}, and then change the AST accordingly.
+There are two main variants to the Abstract Syntax Tree pattern for refactoring:
+mutable and immutable. A mutable tree has nodes that can be easily added, 
+changed, or deleted from the tree. These trees are often used internally by
+compilers. For example, a compiler might determine that
+\ttt{a~=~b*1} can be simplified to \ttt{a~=~b}, and then change the AST 
+accordingly.
 
@@
 
-Unfortunately, while a mutable AST can provide an elegant means of manipulating
-source code, that are other issues that can make mutable ASTs problematic from
-an API design perspective.  Some notable issues include the following.
+Mutable ASTs are conceptually simple and allow code for specific refactorings
+to be very short. Performing refactorings on source code is conceptually
+identical to changing the AST itself. This can greatly lower the bar for 
+designing refactorings. However, mutable ASTs introduce a host of problems when
+used in a large API:
+
 \begin{itemize}
@@
 emphasize that APIs should be easy to learn (even for programmers who do not
-read all of the API documentation) and difficult for clients to misuse.  Given
-these guidelines, and the list of liabilities for mutable ASTs given above, it
-seems preferable from the perspective of someone attempting to publish an API
-to avoid mutable ASTs.  But this raises a different question: If ASTs are
-immutable, what API should be provided for clients to manipulate source code?
+read all of the API documentation) and difficult for clients to misuse. Mutable
+ASTs have the advantage of ease-of-learning, but have have all the disadvantages
+listed above. They can be valid solutions for simpler refactorings or projects
+developed by a small team of people, but become problematic when used for 
+nontrivial refactorings or for a public API. For a general, public refactoring
+API, immutable trees are less error-prone.
 
@@
 So, is it possible to provide a cleaner source manipulation API, while still
-using an immutable AST?  It is, but some additional API is needed.
+using an immutable AST?  It is, and there are two main variants that allow
+for this.
 
-One option is to provide an API that allows clients to specify what changes
-they want to make in terms of AST nodes.  This API can then translate these
-into string edits.  For example, an API might offer a \textit{Rewriter} class.
-The client would inform the \textit{Rewriter} that it wants to effectively
-delete a particular \textit{FieldDeclaration} node from the AST:
+\subsection{Variants}
+
+\begin{itemize}
+
+    \item \textit{Work with the ASTs in a functional style, where every edit
+        returns a new AST.} So, a call to field.removeFromTree() would not
+        change the existing AST, but rather it would return a new AST that has
+        the given field omitted. When dealing with immutable data structures,
+        functional programming may often be the natural solution. However,
+        in the case of ASTs, a functional style risks reintroducing many of 
+        the problems inherent to mutable trees. File location of nodes may
+        be disrupted, pointers to nodes in the original tree may be rendered 
+        irrelevant, semantic information may be invalidated, etc. Those are in
+        addition to the difficulties of functional programming, such as memory
+        consumption and copying, especially with very large ASTs. This is
+        a valid refactoring API design pattern, but any implementation using 
+        this variant must carefully consider these problems.
+
+    \item \textit{Provide an API that allows clients to specify what changes
+        they want to make in terms of AST nodes.}  This API can then translate these into string edits.  For example, an API might offer a \textit{Rewriter} 
+class. The client would inform the \textit{Rewriter} that it wants to 
+effectively delete a particular \textit{FieldDeclaration} node from the AST:
 \textit{rewriter.delete(field)}.  When it is ready to manipulate the source
@@
 
-A second option is to work with ASTs in a functional style, where every edit
-returns a new AST.  So, a call to \textit{field.removeFromTree()} would not
-change the existing AST, but rather it would return a new AST that has the
-given field omitted.  (Of course, this requires a very careful implementation.
-Memory consumption and copying are a concern, particularly when ASTs are large.
-Also, a mechanism must be provided so that clients can easily map nodes in the
-original tree to the ``same'' node in the modified tree, and vice versa.)
+\end{itemize}
 

### 2016-07-20 paper.tex
The example in section 4 now uses the same refactoring as sectioin 2.

@@
 \documentclass[prodmode]{acmlarge}
-%\documentclass[draft]{acmlarge}
+%\documentclass[draft,prodmode]{acmlarge}
 \usepackage{graphicx}
@@
 nontrivial refactorings or for a public API. For a general, public refactoring
-API, immutable trees are less error-prone.
+API, immutable trees are much more reliable.
 
@@
 The situation gets a bit worse when changes are made to nested constructs.
-Recall the example from Section~\ref{sec:Problem}, which added braces to two
-\textit{if} statements, one nested inside the other.  Suppose the outer
-\textit{if} statement were modified first: Adding braces to it would change the
-position at which the closing brace for the inner \textit{if} statement would
-need to be added.  Consider a somewhat simplified example.  Pictorially:
+Recall the example from Section~\ref{sec:Problem}, which reversed conditionals 
+of \textit{if} statements. After the addition of the \textit{!} to the 
+condition, the offset of the body is different. Consider a somewhat simplified 
+example. Pictorially:
 
@@
 \jb{i}\jb{f}\jb{(}\jb{1}\jb{)}\jb{ }%
-\jb{i}\jb{f}\jb{(}\jb{2}\jb{)}\jb{ }%
+\jb{s}\jb{(}\jb{)}\jb{;}\jb{ }%
+\jb{e}\jb{l}\jb{s}\jb{e}\jb{ }%
+\jb{b}\jb{(}\jb{)}\jb{;}\\
+\lb{\,}\kb{0}\kb{1}\kb{2}\kb{3}\kb{4}\kb{5}\kb{6}\kb{7}\kb{8}\lb{9}%
+\lb{10}\lb{11}\lb{12}\lb{13}\lb{14}\lb{15}\lb{16}\lb{17}\lb{18}\lb{19}\\
+$\Downarrow$ \rule[-0.5em]{0pt}{1.2em} \\
+\jb{i}\jb{f}\jb{(}\jb{!}\jb{1}\jb{)}\jb{ }%
+\jb{b}\jb{(}\jb{)}\jb{;}\jb{ }%
+\jb{e}\jb{l}\jb{s}\jb{e}\jb{ }%
 \jb{s}\jb{(}\jb{)}\jb{;}\\
 \lb{\,}\kb{0}\kb{1}\kb{2}\kb{3}\kb{4}\kb{5}\kb{6}\kb{7}\kb{8}\lb{9}%
-\lb{10}\lb{11}\lb{12}\lb{13}\lb{14}\lb{15} \\
-$\Downarrow$ \rule[-0.5em]{0pt}{1.2em} \\
-\jb{i}\jb{f}\jb{(}\jb{1}\jb{)}\jb{ }%
-\jb{\{}\jb{ }%
-\jb{i}\jb{f}\jb{(}\jb{2}\jb{)}\jb{ }%
-\jb{s}\jb{(}\jb{)}\jb{;}%
-\jb{ }\jb{\}}\\
-\lb{\,}\kb{0}\kb{1}\kb{2}\kb{3}\kb{4}\kb{5}\kb{6}\kb{7}\kb{8}\lb{9}%
-\lb{10}\lb{11}\lb{12}\lb{13}\lb{14}\lb{15}\lb{16}\lb{17}\lb{18}\lb{19} \\
+\lb{10}\lb{11}\lb{12}\lb{13}\lb{14}\lb{15}\lb{16}\lb{17}\lb{18}\lb{19}\lb{20}\\
 \end{center}
@@
 \noindent
-The body of the inner \textit{if} statement begins at offset~12 in the original
+The body of the first \textit{if} statement begins at offset~6 in the original
 code.  But after adding braces to the outer \textit{if} statement, the body of
-the inner \textit{if} statement is shifted to offset~14.
+the inner \textit{if} statement is shifted to offset~7. This must be taken into
+account when the bodies of the \textit{if} and \textit{else} are switched.
 

### 2016-07-22 paper.tex
Added basic info on preprocessors, needs more work though.

@@
 
+%TODO work on this preprocessor stuff some more
+%Maybe move to a different place?
+Preprocessors (such as the C preprocessor and M4), don't interfere with the ASTs
+used by compilers. In most cases, the preprocessing is performed first, and the
+output thereof is used to construct the AST. The lexer/parser requires no 
+knowledge of the fact that the text was preprocessed.
+
+Refactorings, however, are complicated by the C preprocessor. The grammar of 
+preprocessor directives doesn't take into account C syntax, and can't be easily
+represented in the C AST. This is the subject of Garrido's entire Ph.D.~thesis~%
+\cite{garrido05program}. The compiler's method of preprocessing before parsing
+can't be used, because the unpreprocessed text is what must be edited. The 
+refactoring too must be able to parse, analyze, and transform code \textit{
+with embedded preprocessor directives.}
+
+The topic of refactoring preprocessed code will not be treated in detail here,
+but suffice it to say that a compiler's AST will almost certainly be inadequate.
+For more details, see Overbey et al.~\cite{wrt2009} and 
+Garrido~\cite{garrido05program}.
+
 %Preprocessors (such as the C preprocessor and M4) do not pose much of a problem
@@
 Recall the example from Section~\ref{sec:Problem}, which reversed conditionals 
-of \textit{if} statements. After the addition of the \textit{!} to the 
+of \textit{if} statements. After the addition of the \textit{not} to the 
 condition, the offset of the body is different. Consider a somewhat simplified 
@@
 The body of the first \textit{if} statement begins at offset~6 in the original
-code.  But after adding braces to the outer \textit{if} statement, the body of
+code.  But after interchanging the body statements, the body of
 the inner \textit{if} statement is shifted to offset~7. This must be taken into

### 2016-07-26 paper.tex
Cleaned up a few spots

@@
 \item The API should be usable when building refactorings for large code bases.
-Automated refactorings are most useful for maintaining such codes.
+Automated refactorings are most useful for maintaining such code.
 
@@
 
-Mutable ASTs are conceptually simple and allow code for specific refactorings
+Mutable ASTs are simple to understand and allow code for specific refactorings
 to be very short. Performing refactorings on source code is conceptually
@@
 nontrivial refactorings or for a public API. For a general, public refactoring
-API, immutable trees are much more reliable.
+API, immutable trees are much more robust.
 
@@
         the problems inherent to mutable trees. File location of nodes may
-        be disrupted, pointers to nodes in the original tree may be rendered 
+        be disrupted in the new tree, pointers to nodes in the original tree ma
+        be rendered 
         irrelevant, semantic information may be invalidated, etc. Those are in
@@
 efficient mid-string insertions and
-deletions\footnote{\url{http://clang.llvm.org/doxygen/classclang_1_1RewriteRope.html}}.
+deletions\footnote{RewriteRope Class Reference: \url{http://clang.llvm.org/doxygen/classclang_1_1RewriteRope.html}}.
 (It uses a B-tree structure inspired by, but not identical to, that described

### 2016-07-29 paper.tex
Moved preprocessor stuff to its own subsection.

@@
 is available for every AST node, it becomes fairly easy to determine what
-syntactic construct corresponds to the text selection corresponds to: simply
+syntactic the text selection corresponds to: simply
 walk the tree, searching for a node whose source location aligns with the
@@
 %Maybe move to a different place?
+\subsection{Preprocessors}
 Preprocessors (such as the C preprocessor and M4), don't interfere with the ASTs
@@
 Garrido~\cite{garrido05program}.
-
-%Preprocessors (such as the C preprocessor and M4) do not pose much of a problem
-%for compilers.  In a compiler, the preprocessor can be run first, and then its
-%output can serve as the input to the lexer/parser.  The preprocessor can be
-%completely unaware of what lexer/parser it is feeding into, and the
-%lexer/parser can be unaware of the fact that the code was preprocessed
-%(although it might handle directives like \code{\#line} to give better error
-%messages).
-%
-%Unfortunately, the C preprocessor makes refactoring very difficult; it is the
-%subject of Garrido's entire Ph.D.~thesis~\cite{garrido05program}.  The
-%compiler's approach of preprocesing the code first and then parsing it does not
-%work.  Since refactoring tools must modify the \textit{user's} source code,
-%they must be able to parse, analyze, and transform code \textit{with embedded
-%preprocessor directives.}
-%
-%This dissertation will not treat the topic of refactoring preprocessed code in
-%detail, but suffice it to say that a compiler's AST will almost certainly be
-%inadequate.  At a minimum, a refactoring must be able to determine which AST
-%nodes correspond to actual text in the user's source code and which ones are
-%the result of file inclusions and macro expansions.  For more details, see
-%Overbey et al.~\cite{wrt2009} and Garrido~\cite{garrido05program}.
-
-%\subsection{Mapping AST Nodes to Source Locations}
-%\label{ss:mapping}
 

### 2016-08-11 paper.tex
The C plugin now works.

@@
 
-\acmformat{Overbey, J. 2013. Immutable Source-Mapped Abstract Syntax Tree.}
+\acmformat{Overbey, J. and Worley, C. 2013. Immutable Source-Mapped Abstract 
+Syntax Tree.}
 
@@
 redundant parentheses. The former is simple enough, but the latter requires 
-knowledge of matching parentheses. Interchanging the branches similarly requires 
-knowledge of the matching braces, the location of the \textit{if} and dangling 
-\textit{else}, and the presence or absence of any statements that alter flow of 
-control such as \textit{goto}. In essence, knowledge of the program's syntax is 
+knowledge of matching parentheses and, in more complex cases, knowledge of operator
+precedence. Interchanging the branches similarly requires 
+knowledge of the matching braces, the location of the \textit{if}, and presence
+or absence of the dangling 
+\textit{else}.  In essence, knowledge of the program's syntax is 
 required.
@@
 
-%TODO work on this preprocessor stuff some more
-%Maybe move to a different place?
-\subsection{Preprocessors}
-Preprocessors (such as the C preprocessor and M4), don't interfere with the ASTs
-used by compilers. In most cases, the preprocessing is performed first, and the
-output thereof is used to construct the AST. The lexer/parser requires no 
-knowledge of the fact that the text was preprocessed.
-
-Refactorings, however, are complicated by the C preprocessor. The grammar of 
-preprocessor directives doesn't take into account C syntax, and can't be easily
-represented in the C AST. This is the subject of Garrido's entire Ph.D.~thesis~%
-\cite{garrido05program}. The compiler's method of preprocessing before parsing
-can't be used, because the unpreprocessed text is what must be edited. The 
-refactoring too must be able to parse, analyze, and transform code \textit{
-with embedded preprocessor directives.}
-
-The topic of refactoring preprocessed code will not be treated in detail here,
-but suffice it to say that a compiler's AST will almost certainly be inadequate.
-For more details, see Overbey et al.~\cite{wrt2009} and 
-Garrido~\cite{garrido05program}.
-
 \subsection{Mutable vs.\ Immutable ASTs}
@@
         the problems inherent to mutable trees. File location of nodes may
-        be disrupted in the new tree, pointers to nodes in the original tree ma
+        be disrupted in the new tree, pointers to nodes in the original tree may
         be rendered 
@@
 
-%\section{Example}
+%TODO work on this preprocessor stuff some more
+%Maybe move to a different place?
+\subsection{Preprocessors}
+Preprocessors (such as the C preprocessor and M4), don't interfere with the ASTs
+used by compilers. In most cases, the preprocessing is performed first, and the
+output thereof is used to construct the AST. The lexer/parser requires no 
+knowledge of the fact that the text was preprocessed.
+
+Refactorings, however, are complicated by the C preprocessor. The grammar of 
+preprocessor directives doesn't take into account C syntax, and can't be easily
+represented in the C AST. This is the subject of Garrido's entire Ph.D.~thesis~%
+\cite{garrido05program}. The compiler's method of preprocessing before parsing
+can't be used, because the unpreprocessed text is what must be edited. The 
+refactoring too must be able to parse, analyze, and transform code \textit{
+with embedded preprocessor directives.}
+
+The topic of refactoring preprocessed code will not be treated in detail here,
+but suffice it to say that a compiler's AST will almost certainly be inadequate.
+For more details, see Overbey et al.~\cite{wrt2009} and 
+Garrido~\cite{garrido05program}.
 

### 2016-08-11 Reverse Conditional /src/reverse/conditional/c/ReverseConditionalCAlteration.java

@@ -1,76 +1,80 @@
 package reverse.conditional.c;
 
+import org.eclipse.cdt.core.dom.ast.ASTNodeFactoryFactory;
+import org.eclipse.cdt.core.dom.ast.IASTExpression;
+import org.eclipse.cdt.core.dom.ast.IASTNode;
+import org.eclipse.cdt.core.dom.ast.IASTStatement;
+import org.eclipse.cdt.core.dom.ast.IASTTranslationUnit;
+import org.eclipse.cdt.core.dom.ast.IASTUnaryExpression;
+import org.eclipse.cdt.core.dom.ast.INodeFactory;
+import org.eclipse.cdt.core.dom.ast.c.ICNodeFactory;
 import org.eclipse.jdt.core.JavaModelException;
 import org.eclipse.jdt.core.dom.AST;
 import org.eclipse.jdt.core.dom.ASTNode;
-import org.eclipse.jdt.core.dom.CompilationUnit;
-import org.eclipse.jdt.core.dom.Expression;
+import org.eclipse.cdt.core.dom.ast.IASTIfStatement;
 import org.eclipse.jdt.core.dom.IfStatement;
 import org.eclipse.jdt.core.dom.NodeFinder;
 import org.eclipse.jdt.core.dom.ParenthesizedExpression;
 import org.eclipse.jdt.core.dom.PrefixExpression;
-import org.eclipse.jdt.core.dom.Statement;
 import org.eclipse.jdt.core.dom.PrefixExpression.Operator;
-import org.eclipse.jdt.core.dom.rewrite.ASTRewrite;
+import org.eclipse.jdt.core.dom.Statement;
+import org.eclipse.cdt.core.dom.rewrite.ASTRewrite;
 
 public class ReverseConditionalCAlteration {
 
-	private CompilationUnit compNode;
-	private int offset, length;
+	private IASTTranslationUnit AST;
+	private IASTNode selectedNode;
 	
-	public ReverseConditionalCAlteration(CompilationUnit compNode, int offset, int length) {
-		this.compNode = compNode;
-		this.offset = offset;
-		this.length = length;
+	public ReverseConditionalCAlteration(IASTTranslationUnit AST, IASTNode selectedNode) {
+		this.AST = AST;
+		this.selectedNode = selectedNode;
 	}
-	
+
 	public void change(ASTRewrite rewriter) throws JavaModelException {
-		NodeFinder finder = new NodeFinder(this.compNode, this.offset, this.length);
[…]

### 2016-08-11 Reverse Conditional /src/reverse/conditional/c/ReverseConditionalCHandler.java

@@ -57,6 +57,7 @@ public class ReverseConditionalCHandler extends AbstractHandler {
 		int length = ts.getLength();
 		IASTNodeSelector finder = AST.getNodeSelector(null);
 		IASTNode selected = finder.findEnclosingNode(offset, length);
+		
 		return new ReverseConditionalCRefactoring(AST, selected, offset, length);
 	}
 }

### 2016-08-11 Reverse Conditional /src/reverse/conditional/c/ReverseConditionalCRefactoring.java

@@ -1,31 +1,23 @@
 package reverse.conditional.c;
 
+import org.eclipse.cdt.core.dom.ast.IASTCompoundStatement;
 import org.eclipse.cdt.core.dom.ast.IASTIfStatement;
 import org.eclipse.cdt.core.dom.ast.IASTNode;
+import org.eclipse.cdt.core.dom.ast.IASTStatement;
 import org.eclipse.cdt.core.dom.ast.IASTTranslationUnit;
+import org.eclipse.cdt.core.dom.rewrite.ASTRewrite;
 import org.eclipse.core.runtime.CoreException;
 import org.eclipse.core.runtime.IProgressMonitor;
 import org.eclipse.core.runtime.OperationCanceledException;
-import org.eclipse.jdt.core.ICompilationUnit;
-import org.eclipse.jdt.core.dom.AST;
-import org.eclipse.jdt.core.dom.ASTNode;
-import org.eclipse.jdt.core.dom.ASTParser;
-import org.eclipse.jdt.core.dom.CompilationUnit;
-import org.eclipse.jdt.core.dom.IfStatement;
-import org.eclipse.jdt.core.dom.rewrite.ASTRewrite;
-import org.eclipse.jdt.core.refactoring.CompilationUnitChange;
 import org.eclipse.ltk.core.refactoring.Change;
 import org.eclipse.ltk.core.refactoring.RefactoringStatus;
-import org.eclipse.text.edits.MultiTextEdit;
 
 import reverse.conditional.ReverseConditionalRefactoring;
-import reverse.conditional.java.ReverseConditionalJavaAlteration;
 
 public class ReverseConditionalCRefactoring extends ReverseConditionalRefactoring {
 
 	private IASTTranslationUnit AST;
 	private IASTNode selectedNode;
-	private CompilationUnit compNode;
 	private int offset, length;
 	
 	ReverseConditionalCRefactoring() {}
@@ -33,6 +25,8 @@ public class ReverseConditionalCRefactoring extends ReverseConditionalRefactorin
 	ReverseConditionalCRefactoring(IASTTranslationUnit AST, IASTNode selected, int offset, 
 			int length) {
 		this.AST = AST;
+		this.selectedNode = selected;
+		
 		this.selectedNode = selected;
 		this.offset = offset;
 		this.length = length;
@@ -49,13 +43,7 @@ public class ReverseConditionalCRefactoring extends ReverseConditionalRefactorin
[…]

### 2016-08-11 Reverse Conditional /src/reverse/conditional/java/ReverseConditionalJavaAlteration.java
Java plugin now accepts imperfect refactorings.

@@ -3,6 +3,7 @@ package reverse.conditional.java;
 import org.eclipse.jdt.core.JavaModelException;
 import org.eclipse.jdt.core.dom.AST;
 import org.eclipse.jdt.core.dom.ASTNode;
+import org.eclipse.jdt.core.dom.Block;
 import org.eclipse.jdt.core.dom.CompilationUnit;
 import org.eclipse.jdt.core.dom.Expression;
 import org.eclipse.jdt.core.dom.IfStatement;
@@ -26,12 +27,20 @@ public class ReverseConditionalJavaAlteration {
 	
 	public void change(ASTRewrite rewriter) throws JavaModelException {
 		NodeFinder finder = new NodeFinder(this.compNode, this.offset, this.length);
-		IfStatement selectedNode = (IfStatement) finder.getCoveringNode();
+		ASTNode selectedNode = finder.getCoveringNode();
+		if (selectedNode instanceof Block) {
+			for (Object child : ((Block) selectedNode).statements()) {
+				if (((ASTNode) child).getNodeType() == ASTNode.IF_STATEMENT) {
+					selectedNode = (ASTNode) child;
+					break;
+				}
+			}
+		}
 		AST ast = this.compNode.getAST();
 		
-		Expression ifExpression = selectedNode.getExpression();
+		Expression ifExpression = ((IfStatement) selectedNode).getExpression();
 		Expression copiedNode = (Expression) ASTNode.copySubtree(ast, ifExpression);
-		swapStatements(rewriter, selectedNode);
+		swapStatements(rewriter, (IfStatement) selectedNode);
 		
 		negateIfExpression(rewriter, ast, ifExpression, copiedNode);
 	}

### 2016-08-11 Reverse Conditional /src/reverse/conditional/java/ReverseConditionalJavaRefactoring.java

@@ -7,8 +7,10 @@ import org.eclipse.jdt.core.ICompilationUnit;
 import org.eclipse.jdt.core.dom.AST;
 import org.eclipse.jdt.core.dom.ASTNode;
 import org.eclipse.jdt.core.dom.ASTParser;
+import org.eclipse.jdt.core.dom.Block;
 import org.eclipse.jdt.core.dom.CompilationUnit;
 import org.eclipse.jdt.core.dom.IfStatement;
+import org.eclipse.jdt.core.dom.Statement;
 import org.eclipse.jdt.core.dom.rewrite.ASTRewrite;
 import org.eclipse.jdt.core.refactoring.CompilationUnitChange;
 import org.eclipse.ltk.core.refactoring.Change;
@@ -53,7 +55,7 @@ public class ReverseConditionalJavaRefactoring extends ReverseConditionalRefacto
             parser.setBindingsRecovery(true);
             this.compNode = (CompilationUnit) parser.createAST(pm);
 		} else {
-			status.addFatalError("Not a valid selection.");
+			status.addFatalError("Not a valid selection. Selected node is " + this.selectedNode.getClass().toString());
 		}
 		pm.done();
 		
@@ -61,6 +63,14 @@ public class ReverseConditionalJavaRefactoring extends ReverseConditionalRefacto
 	}
 	
 	private boolean validNode() {
+		if (this.selectedNode instanceof Block) {
+			for (Object child : ((Block) this.selectedNode).statements()) {
+				if (((ASTNode) child).getNodeType() == ASTNode.IF_STATEMENT) {
+					this.selectedNode = (ASTNode) child;
+					break;
+				}
+			}
+		}
 		if (this.selectedNode.getNodeType() != ASTNode.IF_STATEMENT) {
 			return false;
 		} else if (((IfStatement) this.selectedNode).getElseStatement() == null) {

### 2016-08-11 paper.tex
Short section talking about the example plugin.

@@
 % Title portion
-\title{Immutable Source-Mapped Abstract Syntax Tree:\ \\
+\title{Immutable Source-Mapped Abstract Syntax Tree:\ \, \\
  A Design Pattern for Refactoring Engine APIs} % \\ \today}
@@
 
-\acmformat{Overbey, J. and Worley, C. 2013. Immutable Source-Mapped Abstract 
+\acmformat{Overbey, J. and Worley, C. 2016. Immutable Source-Mapped Abstract 
 Syntax Tree.}
@@
 
-There are two main variants to the Abstract Syntax Tree pattern for refactoring:
+There are two main variants of abstract syntax tree:
 mutable and immutable. A mutable tree has nodes that can be easily added, 
@@
 code, it delegates this task to the rewriter:
-\textit{rewriter.rewrite(sourceCode)}.  This is illustrated in
+\textit{rewriter.rewrite(sourceCode)}.  This allows multiple edits to be made 
+without conflicting with each other. This is illustrated in
 Figure~\ref{fig:seq-diagram}.  The Eclipse Java and C/C++ Development
@@
 but suffice it to say that a compiler's AST will almost certainly be inadequate.
-For more details, see Overbey et al.~\cite{wrt2009} and 
-Garrido~\cite{garrido05program}.
+For more details, see Garrido~\cite{garrido2013} and Overbey et al.~\cite{wrt2009}.
 
@@
 
-%\section{Related Patterns and Guidelines}
+\section{Example}
+We created a plugin for Eclipse that exhibits the \emph{Rewriter} variant of this
+pattern. It performs the \emph{reverse conditional} refactoring described 
+earlier in both JDT and CDT. The pattern is most visible in the plugin's
+swapStatements method, Figure~\ref{fig:swapStatements}, which swaps the 
+branches of the \emph{if} statement.
+
+As can be seen, the AST provides simple methods to retrieve individual nodes as
+objects. The desired changes are then described to the \emph{rewriter}, which
+doesn't make any actual changes until all edits are collected. In the case of
+this refactoring, the only other change is the negation of the conditional 
+expression, which is performed in a similar manner, replacing the existing
+expression with a negated one.
+
+The pattern is similar in both CDT and JDT, to the extent that the plugin is
+structured identically, with slightly different class names in each.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.30]{swapStatements.eps}
+\caption{The \emph{swapStatements} method.\vspace*{-4em}}
+\label{fig:swapStatements}
+\end{center}
+\end{figure}
 

### 2017-09-26 paper.tex
Fixed ACM Reference Format month/year
Added github repo
Variant explicit names

@@
 \articleSeq{0}
-\acmYear{2013}
-\acmMonth{5}
+\acmYear{2016}
+\acmMonth{8}
 
@@
 
-    \item \textit{Work with the ASTs in a functional style, where every edit
+    \item Functional: \textit{Work with the ASTs in a functional style, where every edit
         returns a new AST.} So, a call to field.removeFromTree() would not
@@
 
-    \item \textit{Provide an API that allows clients to specify what changes
+    \item Rewriter: \textit{Provide an API that allows clients to specify what changes
         they want to make in terms of AST nodes.}  This API can then translate these into string edits.  For example, an API might offer a \textit{Rewriter} 
@@
 We created a plugin for Eclipse that exhibits the \emph{Rewriter} variant of this
-pattern. It performs the \emph{reverse conditional} refactoring described 
+pattern\footnote{GitHub Repository: \url{https://github.com/aggrand/reverse-conditional}}. It performs the \emph{reverse conditional} refactoring described 
 earlier in both JDT and CDT. The pattern is most visible in the plugin's

### 2017-09-27 paper.tex
Made variants a subsection again

@@
 
+\subsection{Basic Forces}
+
+There are several basic forces at work in this problem:
+
 \begin{itemize}
@@
 
-\section{Solution}
-
-\textit{Use a parser for the language being refactored to construct an
-\textup{abstract syntax tree}, or \textup{AST}.  Make the AST
-\textup{immutable}---i.e., its structure cannot be changed---but provide
-\textup{source mapping} information so that every node in the AST can be mapped
-to a particular substring of the original source code.}
+Traditionally, program syntax is made available through abstract syntax trees.
 
@@
 the mutable AST is not accessible via any public API.  It is intended for
-internal use only, by developers who should be better equipped to handle all of
+internal use only, by those tools' developers who should be better equipped to handle all of
 the caveats.
@@
 API, immutable trees are much more robust.
+
+%TODO work on this preprocessor stuff some more
+%Maybe move to a different place?
+\subsection{Preprocessors}
+Preprocessors (such as the C preprocessor and M4), don't interfere with the ASTs
+used by compilers. In most cases, the preprocessing is performed first, and the
+output thereof is used to construct the AST. The lexer/parser requires no 
+knowledge of the fact that the text was preprocessed.
+
+Refactorings, however, are complicated by the C preprocessor. The grammar of 
+preprocessor directives doesn't take into account C syntax, and can't be easily
+represented in the C AST. This is the subject of Garrido's entire Ph.D.~thesis~%
+\cite{garrido05program}. The compiler's method of preprocessing before parsing
+can't be used, because the unpreprocessed text is what must be edited. The 
+refactoring too must be able to parse, analyze, and transform code \textit{
+with embedded preprocessor directives.}
+
+The topic of refactoring preprocessed code will not be treated in detail here,
+but suffice it to say that a compiler's AST will almost certainly be inadequate.
+For more details, see Garrido~\cite{garrido2013} and Overbey et al.~\cite{wrt2009}.
+
+\section{Solution}
+
+\textit{Use a parser for the language being refactored to construct an
+\textup{abstract syntax tree}, or \textup{AST}.  Make the AST
+\textup{immutable}---i.e., its structure cannot be changed---but provide
+\textup{source mapping} information so that every node in the AST can be mapped
+to a particular substring of the original source code.}
 
@@
 
-%TODO work on this preprocessor stuff some more
-%Maybe move to a different place?
-\subsection{Preprocessors}
-Preprocessors (such as the C preprocessor and M4), don't interfere with the ASTs
-used by compilers. In most cases, the preprocessing is performed first, and the
-output thereof is used to construct the AST. The lexer/parser requires no 
-knowledge of the fact that the text was preprocessed.
-
-Refactorings, however, are complicated by the C preprocessor. The grammar of 
-preprocessor directives doesn't take into account C syntax, and can't be easily
-represented in the C AST. This is the subject of Garrido's entire Ph.D.~thesis~%
-\cite{garrido05program}. The compiler's method of preprocessing before parsing
-can't be used, because the unpreprocessed text is what must be edited. The 
-refactoring too must be able to parse, analyze, and transform code \textit{
-with embedded preprocessor directives.}
-
-The topic of refactoring preprocessed code will not be treated in detail here,
-but suffice it to say that a compiler's AST will almost certainly be inadequate.
-For more details, see Garrido~\cite{garrido2013} and Overbey et al.~\cite{wrt2009}.
-
 \section{Known Uses}

### 2017-09-27 paper.tex
Promoted variants back to top-level section
Moved source-mapping solution to the solution section
Moved immutability argument to solution section
Lengthened italics solution text

@@
 
-The easiest way to map AST nodes to source locations is to store source
-location information in the AST nodes themselves.  Again, consider the JDT AST
-shown in Figure~\ref{fig:jdt-ast}, and note the numbers in square brackets.
-The highlighted node is displayed as ``VariableDeclarationFragment [35, 8]'':
-this AST node corresponds to the eight-character string beginning at offset 35
-(\ttt{FIVE = 5}).  Its child node corresponding to the identifier \ttt{FIVE} is
-attributed with the offset-length pair [35, 4], while the numeric literal
-\ttt{5}'s source location is [42, 1].
-
-Now, consider what happens when a user selects text in the text editor.  This
-can be represented as an offset-length pair.  When source location information
-is available for every AST node, it becomes fairly easy to determine what
-syntactic the text selection corresponds to: simply
-walk the tree, searching for a node whose source location aligns with the
-user's selection.
-
-Since AST nodes at deeper levels of the tree represent finer-grained syntactic
-constructs than their ancestor nodes, ASTs with source location information
-typically exhibit a nice property: the source location of a child node will
-always be contained within the source location of its parent.  This property
-can be used to expedite searches based on source location information.  For
-example, if a particular node corresponds to the source location [35, 8], and
-you are searching offset 63, then the desired node cannot possibly be a
-descendent of that node.
-
 \subsection{Mutable vs.\ Immutable ASTs}
@@
 
-Mutable ASTs were used in the original Smalltalk Refactoring
-Browser~\cite{roberts97refactoring}, as well as in
-CRefactory~\cite{garrido05program} and Photran~\cite{sle2008}.  Many compilers
-use mutable ASTs internally.  However, these tools have one thing in common:
-the mutable AST is not accessible via any public API.  It is intended for
-internal use only, by those tools' developers who should be better equipped to handle all of
-the caveats.
-
-Guidelines for API design are discussed in detail in Jaroslav Tulach's
-\textit{Practical API Design}~\cite{tulach08practical}.  A presentation by
-Joshua Bloch~\cite{bloch} offers excellent advice as well.  Both authors
-suggest minimizing the amount of mutability in an API.  Immutable objects tend
-to have simpler interfaces, and they are thread-safe.  Both authors also
-emphasize that APIs should be easy to learn (even for programmers who do not
-read all of the API documentation) and difficult for clients to misuse. Mutable
-ASTs have the advantage of ease-of-learning, but have have all the disadvantages
-listed above. They can be valid solutions for simpler refactorings or projects
-developed by a small team of people, but become problematic when used for 
-nontrivial refactorings or for a public API. For a general, public refactoring
-API, immutable trees are much more robust.
-
 %TODO work on this preprocessor stuff some more
@@
 \textup{source mapping} information so that every node in the AST can be mapped
-to a particular substring of the original source code.}
+to a particular substring of the original source code. Treat the source code
+as a string, and use the source code location information from the AST to
+add, replace, and delete substrings.}
+
+\subsection{Immutability}
+As seen in Section 3.4, mutability introduces a host of problems to ASTs. Mutable ASTs were used in the original Smalltalk Refactoring
+Browser~\cite{roberts97refactoring}, as well as in
+CRefactory~\cite{garrido05program} and Photran~\cite{sle2008}.  Many compilers
+use mutable ASTs internally.  However, these tools have one thing in common:
+the mutable AST is not accessible via any public API.  It is intended for
+internal use only, by those tools' developers who should be better equipped to handle all of
+the caveats.
+
+Guidelines for API design are discussed in detail in Jaroslav Tulach's
+\textit{Practical API Design}~\cite{tulach08practical}.  A presentation by
+Joshua Bloch~\cite{bloch} offers excellent advice as well.  Both authors
+suggest minimizing the amount of mutability in an API.  Immutable objects tend
+to have simpler interfaces, and they are thread-safe.  Both authors also
+emphasize that APIs should be easy to learn (even for programmers who do not
+read all of the API documentation) and difficult for clients to misuse. Mutable
+ASTs have the advantage of ease-of-learning, but have have all the disadvantages
+listed above. They can be valid solutions for simpler refactorings or projects
+developed by a small team of people, but become problematic when used for 
+nontrivial refactorings or for a public API. For a general, public refactoring
+API, immutable trees are much more robust.
+
+\subsection{Source Mapping}
+
+As discussed in Section 3.3, it must be possible to map nodes in the AST to
+locations in the source code, and vice versa. The easiest way to map AST nodes to source locations is to store source
+location information in the AST nodes themselves.  Again, consider the JDT AST
+shown in Figure~\ref{fig:jdt-ast}, and note the numbers in square brackets.
+The highlighted node is displayed as ``VariableDeclarationFragment [35, 8]'':
+this AST node corresponds to the eight-character string beginning at offset 35
+(\ttt{FIVE = 5}).  Its child node corresponding to the identifier \ttt{FIVE} is
+attributed with the offset-length pair [35, 4], while the numeric literal
+\ttt{5}'s source location is [42, 1].
+
+Now, consider what happens when a user selects text in the text editor.  This
+can be represented as an offset-length pair.  When source location information
+is available for every AST node, it becomes fairly easy to determine what
+syntactic the text selection corresponds to: simply
+walk the tree, searching for a node whose source location aligns with the
+user's selection.
+
+Since AST nodes at deeper levels of the tree represent finer-grained syntactic
+constructs than their ancestor nodes, ASTs with source location information
+typically exhibit a nice property: the source location of a child node will
+always be contained within the source location of its parent.  This property
+can be used to expedite searches based on source location information.  For
+example, if a particular node corresponds to the source location [35, 8], and
+you are searching offset 63, then the desired node cannot possibly be a
+descendent of that node.
 
@@
 
-\subsection{Variants}
+\section{Variants}
 

### 2017-09-27 paper.tex
Reworded basic forces section
Basic forces grammar
Removed figure periods

@@
 
-There are several basic forces at work in this problem:
+There are several basic requirements that a solution API must meet.
 
@@
 
-Traditionally, program syntax is made available through abstract syntax trees.
+Traditionally, fine-grained program syntax is made available through 
+abstract syntax trees.
 
@@
 \caption{Visualization of an AST produced by the Eclipse Java Development Tools
-(JDT).\vspace*{-4em}}
+(JDT)\vspace*{-4em}}
 \label{fig:jdt-ast}
@@
 
-\subsection{Mutable vs.\ Immutable ASTs}
+\subsection{Mutability and ASTs}
 \label{ss:manip}
@@
 \end{center}
-\caption{Sequence diagram illustrating use of a Rewriter to modify source code.}
+\caption{Sequence diagram illustrating use of a Rewriter to modify source code}
 \label{fig:seq-diagram}
@@
 \includegraphics[scale=0.30]{swapStatements.eps}
-\caption{The \emph{swapStatements} method.\vspace*{-4em}}
+\caption{The \emph{swapStatements} method\vspace*{-4em}}
 \label{fig:swapStatements}

### 2017-09-27 paper.tex
made solution bold

@@
 
-Traditionally, fine-grained program syntax is made available through 
+Typically, fine-grained program syntax is made available through 
 abstract syntax trees.
@@
 can't be used, because the unpreprocessed text is what must be edited. The 
-refactoring too must be able to parse, analyze, and transform code \textit{
-with embedded preprocessor directives.}
+refactoring too must be able to parse, analyze, and transform code \textit{with embedded preprocessor directives.}
 
@@
 
-\textit{Use a parser for the language being refactored to construct an
-\textup{abstract syntax tree}, or \textup{AST}.  Make the AST
-\textup{immutable}---i.e., its structure cannot be changed---but provide
-\textup{source mapping} information so that every node in the AST can be mapped
+\textbf{Use a parser for the language being refactored to construct an
+\textit{abstract syntax tree}, or \textup{AST}.  Make the AST
+\textit{immutable}---i.e., its structure cannot be changed---but provide
+\textit{source mapping} information so that every node in the AST can be mapped
 to a particular substring of the original source code. Treat the source code

### 2017-09-27 paper.tex
Added a statement of application
Clarified rewriter

@@
 
+Abstract syntax trees have many applications, but these two restrictions are
+unique to refactorings and require a pattern that specifically addresses them.
+
 \subsection{Mutability and ASTs}
@@
 effectively delete a particular \textit{FieldDeclaration} node from the AST:
-\textit{rewriter.delete(field)}.  When it is ready to manipulate the source
+\textit{rewriter.delete(field)}.  The collection of planned edits can be 
+        maintained by the rewriter, which handles all source-AST mapping concerns. When the client is ready to manipulate the source
 code, it delegates this task to the rewriter:

### 2017-09-27 paper.tex
Removed flow-disrupting parentheses

@@
 
-Developers familiar with the HTML or XML DOM (Document Object Model) will find
-this structure familiar.  The DOM is effectively an abstract syntax tree for
-HTML/XML documents.
-
-Deeper discussions on ASTs can be found in many introductory textbooks on
-compiler construction, such as \cite{aho06compilers} and
-\cite{torczon11engineering}.  Such books also discuss how parsers are
-constructed, including how they can be modified to construct ASTs.
-
 \FloatBarrier
@@
 statements as \textsc{write} statements in its AST.  The GNU Fortran compiler
-does this, for example.  However, the ability to convert \textsc{print}
+does this, for example.  However, onsider a program that uses \textsc{print} 
+statements to write to standard output, but the programmer later needs to 
+use \textsc{write}
+statements to write to a file instead. Then the ability to convert \textsc{print}
 statements to \textsc{write} statements actually makes for a useful
-refactoring.  (Consider a program that uses \textsc{print} statements to write
-to standard output, but the programmer later needs to use \textsc{write}
-statements to write to a file instead.) To implement this refactoring, a
+refactoring. To implement this refactoring, a
 refactoring tool needs to be able to distinguish \textsc{print} statements from
@@
 deletions\footnote{RewriteRope Class Reference: \url{http://clang.llvm.org/doxygen/classclang_1_1RewriteRope.html}}.
-(It uses a B-tree structure inspired by, but not identical to, that described
-in the original paper on Ropes~\cite{boehm95ropes}.)
+It uses a B-tree structure inspired by, but not identical to, that described
+in the original paper on Ropes~\cite{boehm95ropes}.
 

### 2017-10-03 paper.tex
Made forces more bulleted

@@
 knowledge of matching parentheses and, in more complex cases, knowledge of operator
+
 precedence. Interchanging the branches similarly requires 
@@
 
-\section{Forces}
-
-\subsection{Basic Forces}
-
-There are several basic requirements that a solution API must meet.
-
-\begin{itemize}
-
-\item The API must expose fine-grained details about the syntax of a program.
-For example, the preceding example requires being able to determine the
-exact location of parentheses \ttt{(} \ttt{)}.
-
-\item Many refactorings are initiated only after the user selects a range of
-source code in a text editor.  Therefore, the API must allow clients to
-determine what syntactic construct(s) are selected.
-
-\item Clients must be able to use the API to manipulate source code in a way
-that is minimally invasive: refactorings that only affect a small, isolated
-region of the source code should not change any source code outside that
-region.  This includes preserving the program's comments and
-formatting~\cite{sommerlad08retaining}.
-
-\item The API must follow general principles of good API design, including
-being easy to learn and difficult to misuse.
-
-\item The API should be usable when building refactorings for large code bases.
-Automated refactorings are most useful for maintaining such code.
-
-\end{itemize}
-
-Typically, fine-grained program syntax is made available through 
-abstract syntax trees.
-
-\subsection{Abstract Syntax Trees}
+\section{Background: Abstract Syntax Trees}
 \label{ss:asts}
@@
 source file: one child node describes the
+
+\noindent
+\ttt{package} statement, another
+contains a list of \ttt{import} statements, and a third contains a list of type
+declarations.  A type declaration has several child nodes, which describe
+(among other things) its visibility, the name of the class or interface, and
+the field/method declarations in its body.  A field declaration has child nodes
+describing its visibility, type, and so forth.
+
+Figure~\ref{fig:jdt-ast} shows a visualization of an AST for the following
+program:
+
+\begin{center}
+\vspace*{1em}
+\begin{tabular}{ccc}
+\begin{minipage}{2.25in}
+\noindent \ttt{public class Example \{ \\
+\hspace*{1.5em} public int FIVE = 5; \\
+\}
+}
+\end{minipage}
+\end{tabular}
+\vspace*{1em}
+\end{center}
+
+Notice how deeper levels of the tree describe progressively more fine-grained
+syntactic structures.  The root node represents the entire Java source file.
+The highlighted node---a \textit{VariableDeclarationFragment}---represents
+\ttt{FIVE = 5}, while its child nodes represent its constituent parts: the
+identifier \ttt{FIVE} and the numeric literal \ttt{5}.  These are leaf nodes,
+since these tokens (like keywords, identifiers, and numeric literals) represent
+the smallest syntactic units in the language.
+
+There are two main variants of abstract syntax tree:
+mutable and immutable. A mutable tree has nodes that can be easily added, 
+changed, or deleted from the tree. 
+These trees are often used internally by
 
@@
 
-\noindent
-\ttt{package} statement, another
-contains a list of \ttt{import} statements, and a third contains a list of type
-declarations.  A type declaration has several child nodes, which describe
-(among other things) its visibility, the name of the class or interface, and
-the field/method declarations in its body.  A field declaration has child nodes
-describing its visibility, type, and so forth.
+compilers. For example, a compiler might determine that
+\ttt{a~=~b*1} can be simplified to \ttt{a~=~b}, and then change
 
-Figure~\ref{fig:jdt-ast} shows a visualization of an AST for the following
-program:
-\vspace*{.5em}
+the AST accordingly. Immutable trees, on the other hand, cannot be changed via
+function calls. These can be used for read-only purposes or operated on in a
+functional manner.
 
-\noindent \ttt{public class Example \{ \\
-\hspace*{.2em} public int FIVE = 5; \\
-\}
-}
+\section{Forces}
 
-\vspace*{.5em}
-Notice how deeper levels of the tree describe progressively more fine-grained
-syntactic structures.  The root node represents the entire Java source file.
-The highlighted node---a \textit{VariableDeclarationFragment}---represents
-\ttt{FIVE = 5}, while its child nodes represent its constituent parts: the
-identifier \ttt{FIVE} and the numeric literal \ttt{5}.  These are leaf nodes,
-since these tokens (like keywords, identifiers, and numeric literals) represent
-the smallest syntactic units in the language.
+\subsection{Basic Forces}
+
+There are several basic requirements that a solution API must meet.
+
+\begin{itemize}
+
+    \item \textit{The API must expose fine-grained details about the syntax of a program.}
+For example, the preceding example requires being able to determine the
+exact location of parentheses \ttt{(} \ttt{)}.
+
+\item \textit{Many refactorings are initiated only after the user selects a range of
+source code in a text editor.  Therefore, the API must allow clients to
+        determine what syntactic construct(s) are selected.}
+
+    \item \textit{Clients must be able to use the API to manipulate source code in a way
+        that is minimally invasive.} Refactorings that only affect a small, isolated
+region of the source code should not change any source code outside that
+region.  This includes preserving the program's comments and
+formatting~\cite{sommerlad08retaining}.
+
+\item \textit{The API must follow general principles of good API design, including
+    being easy to learn and difficult to misuse.}
+
+\item \textit{The API should be usable when building refactorings for large code bases.}
+Automated refactorings are most useful for maintaining such code.
+
+\end{itemize}
+
+Typically, fine-grained program syntax is made available through 
+abstract syntax trees.
+
+\subsection{AST Forces}
+\label{ss:manip}
+
+Since mutable ASTs are used by compilers, they seem like a reasonable mechanism for refactoring tools to use
+as well.  To remove a field declaration from a program, simply delete its
+\textit{FieldDeclaration} node from the AST (along with all its children).  To
+change the name of \ttt{FIVE}, change the contents of its
+\textit{SimpleNameNode} in the AST.  Of course, some mechanism is needed to
+prettyprint the modified AST, so these changes can be translated back into
+source code, but that is relatively straightforward to implement.
+
+Mutable ASTs have several benefits:
+\begin{itemize}
+    \item \textit{Mutable ASTs are simple to understand.} Performing 
+        refactorings on source code is conceptually identical to changing 
+        the AST itself.
+    \item \textit{Source code for specific refactorings can be very short.} 
+        The code simply has to rearrange the AST.
+
+\end{itemize}
+
+However, mutability introduces several issues:
 
 \FloatBarrier
+
+\begin{itemize}
+\item \textit{AST nodes contain source location information.}  After the
+structure of the AST is changed, what does the source location information in
+each node represent: the ``old'' source location (in the original code) or the
+``new'' location (after it has been modified)?  If it represents the old
+location, what should the source location be when new nodes are inserted into
+the tree?  If it represents the new location, then every time a change is made
+to the tree, source location information must be updated for every node in the
+tree, which may be expensive \dots or else a clever implementation (e.g.,
+computation from relative offsets) must be used instead.
+\item \textit{AST nodes may be removed from the tree, leaving them orphaned.}
+Suppose a client traverses the AST, storing pointers to some ``interesting''
+nodes in a collection.  Then, an ancestor of one of those nodes is removed from
+the AST.  This will leave the client in a difficult-to-debug situation where
+his collection contains pointers to nodes that no longer exist in the AST.
+\item \textit{Semantic information may be invalidated.}  Refactorings often
+need to determine, for a particular use of a variable, what declaration that
+use corresponds to.  For example, in Java, if a field and a local variable were
+both named \ttt{x}, then in the expression \ttt{this.x = x}, the first \ttt{x}
+would refer to the field, while the second would refer to the local variable.
+However, if the local variable declaration were removed, the second \ttt{x}
+would refer to the field instead.  This can be confusing to clients, since
+information collected about the program at one point in time may no longer be
+true after the tree's structure changes.
+\item \textit{It may be possible to mutate the AST so that it does not
+represent a legal program.}  It might even be possible to mutate the AST so
+that it is no longer a tree at all.  In Java, classes can contain other
+classes.  What happens if a client tries to make a class an inner class of
+itself?  (This could turn the tree into a cyclic graph.)  The API for modifying
+the AST's structure could try to prevent such things, but attempting to detect
+and prevent every possible mistake would be difficult and potentially
+expensive.
+\item \textit{Mutability can be problematic in multithreaded code.}
+\end{itemize}
+
+Any candidate solution would have to satisfy the API requirements while avoiding
+the problems associated with mutable ASTs.
+
+%TODO work on this preprocessor stuff some more
+%Maybe move to a different place?
+\subsection{Preprocessors}
+Preprocessors (such as the C preprocessor and M4), don't interfere with the ASTs
+used by compilers. In most cases, the preprocessing is performed first, and the
+output thereof is used to construct the AST. The lexer/parser requires no 
+knowledge of the fact that the text was preprocessed.
+
+Refactorings, however, are complicated by the C preprocessor. The grammar of 
+preprocessor directives doesn't take into account C syntax, and can't be easily
+represented in the C AST. This is the subject of Garrido's entire Ph.D.~thesis~%
+\cite{garrido05program}. The compiler's method of preprocessing before parsing
+can't be used, because the unpreprocessed text is what must be edited. The 
+refactoring too must be able to parse, analyze, and transform code \textit{with embedded preprocessor directives.}
+
+The topic of refactoring preprocessed code will not be treated in detail here,
+but suffice it to say that a compiler's AST will almost certainly be inadequate.
+For more details, see Garrido~\cite{garrido2013} and Overbey et al.~\cite{wrt2009}.
+
+\clearpage
+\section{Solution}
+
+\textbf{Use a parser for the language being refactored to construct an
+\textit{abstract syntax tree}, or \textup{AST}.  Make the AST
+\textit{immutable}---i.e., its structure cannot be changed---but provide
+\textit{source mapping} information so that every node in the AST can be mapped
+to a particular substring of the original source code. Treat the source code
+as a string, and use the source code location information from the AST to
+add, replace, and delete substrings.}
 
@@
 statements as \textsc{write} statements in its AST.  The GNU Fortran compiler
-does this, for example.  However, onsider a program that uses \textsc{print} 
+does this, for example.  However, consider a program that uses \textsc{print} 
 statements to write to standard output, but the programmer later needs to 
@@
 
-\subsection{Mutability and ASTs}
-\label{ss:manip}
-
-There are two main variants of abstract syntax tree:
-mutable and immutable. A mutable tree has nodes that can be easily added, 
-changed, or deleted from the tree. These trees are often used internally by
-compilers. For example, a compiler might determine that
-\ttt{a~=~b*1} can be simplified to \ttt{a~=~b}, and then change the AST 
-accordingly.
-
-At first, this seems like a reasonable mechanism for refactoring tools to use
-as well.  To remove a field declaration from a program, simply delete its
-\textit{FieldDeclaration} node from the AST (along with all its children).  To
-change the name of \ttt{FIVE}, change the contents of its
-\textit{SimpleNameNode} in the AST.  Of course, some mechanism is needed to
-prettyprint the modified AST, so these changes can be translated back into
-source code, but that is relatively straightforward to implement.
-
-Mutable ASTs are simple to understand and allow code for specific refactorings
-to be very short. Performing refactorings on source code is conceptually
-identical to changing the AST itself. This can greatly lower the bar for 
-designing refactorings. However, mutable ASTs introduce a host of problems when
-used in a large API:
-
-\begin{itemize}
-\item \textit{AST nodes contain source location information.}  After the
-structure of the AST is changed, what does the source location information in
-each node represent: the ``old'' source location (in the original code) or the
-``new'' location (after it has been modified)?  If it represents the old
-location, what should the source location be when new nodes are inserted into
-the tree?  If it represents the new location, then every time a change is made
-to the tree, source location information must be updated for every node in the
-tree, which may be expensive \dots or else a clever implementation (e.g.,
-computation from relative offsets) must be used instead.
-\item \textit{AST nodes may be removed from the tree, leaving them orphaned.}
-Suppose a client traverses the AST, storing pointers to some ``interesting''
-nodes in a collection.  Then, an ancestor of one of those nodes is removed from
-the AST.  This will leave the client in a difficult-to-debug situation where
-his collection contains pointers to nodes that no longer exist in the AST.
-\item \textit{Semantic information may be invalidated.}  Refactorings often
-need to determine, for a particular use of a variable, what declaration that
-use corresponds to.  For example, in Java, if a field and a local variable were
-both named \ttt{x}, then in the expression \ttt{this.x = x}, the first \ttt{x}
-would refer to the field, while the second would refer to the local variable.
-However, if the local variable declaration were removed, the second \ttt{x}
-would refer to the field instead.  This can be confusing to clients, since
-information collected about the program at one point in time may no longer be
-true after the tree's structure changes.
-\item \textit{It may be possible to mutate the AST so that it does not
-represent a legal program.}  It might even be possible to mutate the AST so
-that it is no longer a tree at all.  In Java, classes can contain other
-classes.  What happens if a client tries to make a class an inner class of
-itself?  (This could turn the tree into a cyclic graph.)  The API for modifying
-the AST's structure could try to prevent such things, but attempting to detect
-and prevent every possible mistake would be difficult and potentially
-expensive.
-\item \textit{Mutability can be problematic in multithreaded code.}
-\end{itemize}
-
-%TODO work on this preprocessor stuff some more
-%Maybe move to a different place?
-\subsection{Preprocessors}
-Preprocessors (such as the C preprocessor and M4), don't interfere with the ASTs
-used by compilers. In most cases, the preprocessing is performed first, and the
-output thereof is used to construct the AST. The lexer/parser requires no 
-knowledge of the fact that the text was preprocessed.
-
-Refactorings, however, are complicated by the C preprocessor. The grammar of 
-preprocessor directives doesn't take into account C syntax, and can't be easily
-represented in the C AST. This is the subject of Garrido's entire Ph.D.~thesis~%
-\cite{garrido05program}. The compiler's method of preprocessing before parsing
-can't be used, because the unpreprocessed text is what must be edited. The 
-refactoring too must be able to parse, analyze, and transform code \textit{with embedded preprocessor directives.}
-
-The topic of refactoring preprocessed code will not be treated in detail here,
-but suffice it to say that a compiler's AST will almost certainly be inadequate.
-For more details, see Garrido~\cite{garrido2013} and Overbey et al.~\cite{wrt2009}.
-
-\section{Solution}
-
-\textbf{Use a parser for the language being refactored to construct an
-\textit{abstract syntax tree}, or \textup{AST}.  Make the AST
-\textit{immutable}---i.e., its structure cannot be changed---but provide
-\textit{source mapping} information so that every node in the AST can be mapped
-to a particular substring of the original source code. Treat the source code
-as a string, and use the source code location information from the AST to
-add, replace, and delete substrings.}
-
-\subsection{Immutability}
-As seen in Section 3.4, mutability introduces a host of problems to ASTs. Mutable ASTs were used in the original Smalltalk Refactoring
-Browser~\cite{roberts97refactoring}, as well as in
-CRefactory~\cite{garrido05program} and Photran~\cite{sle2008}.  Many compilers
-use mutable ASTs internally.  However, these tools have one thing in common:
-the mutable AST is not accessible via any public API.  It is intended for
-internal use only, by those tools' developers who should be better equipped to handle all of
-the caveats.
-
-Guidelines for API design are discussed in detail in Jaroslav Tulach's
-\textit{Practical API Design}~\cite{tulach08practical}.  A presentation by
-Joshua Bloch~\cite{bloch} offers excellent advice as well.  Both authors
-suggest minimizing the amount of mutability in an API.  Immutable objects tend
-to have simpler interfaces, and they are thread-safe.  Both authors also
-emphasize that APIs should be easy to learn (even for programmers who do not
-read all of the API documentation) and difficult for clients to misuse. Mutable
-ASTs have the advantage of ease-of-learning, but have have all the disadvantages
-listed above. They can be valid solutions for simpler refactorings or projects
-developed by a small team of people, but become problematic when used for 
-nontrivial refactorings or for a public API. For a general, public refactoring
-API, immutable trees are much more robust.
-
 \subsection{Source Mapping}
 
-As discussed in Section 3.3, it must be possible to map nodes in the AST to
-locations in the source code, and vice versa. The easiest way to map AST nodes to source locations is to store source
+The easiest way to map AST nodes to source locations is to store source
 location information in the AST nodes themselves.  Again, consider the JDT AST
@@
 descendent of that node.
+
+\subsection{Immutability}
+
+Mutable ASTs were used in the original Smalltalk Refactoring
+Browser~\cite{roberts97refactoring}, as well as in
+CRefactory~\cite{garrido05program} and Photran~\cite{sle2008}.  Many compilers
+use mutable ASTs internally.  However, these tools have one thing in common:
+the mutable AST is not accessible via any public API.  It is intended for
+internal use only, by those tools' developers who should be better equipped to handle all of
+the caveats.
+
+Guidelines for API design are discussed in detail in Jaroslav Tulach's
+\textit{Practical API Design}~\cite{tulach08practical}.  A presentation by
+Joshua Bloch~\cite{bloch} offers excellent advice as well.  Both authors
+suggest minimizing the amount of mutability in an API.  Immutable objects tend
+to have simpler interfaces, and they are thread-safe.  Both authors also
+emphasize that APIs should be easy to learn (even for programmers who do not
+read all of the API documentation) and difficult for clients to misuse. Mutable
+ASTs have the advantage of ease-of-learning, but have have all the disadvantages
+listed above. They can be valid solutions for simpler refactorings or projects
+developed by a small team of people, but become problematic when used for 
+nontrivial refactorings or for a public API. For a general, public refactoring
+API, immutable trees are much more robust.
 

### 2017-10-04 paper.tex
Almost final touches

@@
 changed, or deleted from the tree. 
-These trees are often used internally by
-
+These trees are often used internally by compilers.
 \begin{wrapfigure}{r}{3in} % r is right, R is float right
@@
 \end{wrapfigure}
-
-compilers. For example, a compiler might determine that
+For example, a compiler might determine that
 \ttt{a~=~b*1} can be simplified to \ttt{a~=~b}, and then change 
-
 the AST accordingly. Immutable trees, on the other hand, cannot be changed via
@@
 
+\label{ss:reqs}
 \subsection{Basic Forces}
 
-There are several basic requirements that a solution API must meet.
+To be useful for implementing refactorings, and API exposed by a refactoring 
+tool must meet several basic requirements:
 
@@
 
-\subsection{AST Forces}
-\label{ss:manip}
+\subsection{Scalability Forces}
 
-Since mutable ASTs are used by compilers, they seem like a reasonable mechanism for refactoring tools to use
-as well.  To remove a field declaration from a program, simply delete its
-\textit{FieldDeclaration} node from the AST (along with all its children).  To
-change the name of \ttt{FIVE}, change the contents of its
-\textit{SimpleNameNode} in the AST.  Of course, some mechanism is needed to
-prettyprint the modified AST, so these changes can be translated back into
-source code, but that is relatively straightforward to implement.
-
-Mutable ASTs have several benefits:
-\begin{itemize}
-    \item \textit{Mutable ASTs are simple to understand.} Performing 
-        refactorings on source code is conceptually identical to changing 
-        the AST itself.
-    \item \textit{Source code for specific refactorings can be very short.} 
-        The code simply has to rearrange the AST.
-
-\end{itemize}
-
-However, mutability introduces several issues:
-
-\FloatBarrier
+As a code base increases in size, so does its complexity. A refactoring API 
+must be able to handle the constraints that a more complicated program 
+introduces:
 
 \begin{itemize}
-\item \textit{AST nodes contain source location information.}  After the
-structure of the AST is changed, what does the source location information in
-each node represent: the ``old'' source location (in the original code) or the
-``new'' location (after it has been modified)?  If it represents the old
-location, what should the source location be when new nodes are inserted into
-the tree?  If it represents the new location, then every time a change is made
-to the tree, source location information must be updated for every node in the
-tree, which may be expensive \dots or else a clever implementation (e.g.,
-computation from relative offsets) must be used instead.
-\item \textit{AST nodes may be removed from the tree, leaving them orphaned.}
-Suppose a client traverses the AST, storing pointers to some ``interesting''
-nodes in a collection.  Then, an ancestor of one of those nodes is removed from
-the AST.  This will leave the client in a difficult-to-debug situation where
-his collection contains pointers to nodes that no longer exist in the AST.
-\item \textit{Semantic information may be invalidated.}  Refactorings often
-need to determine, for a particular use of a variable, what declaration that
-use corresponds to.  For example, in Java, if a field and a local variable were
-both named \ttt{x}, then in the expression \ttt{this.x = x}, the first \ttt{x}
-would refer to the field, while the second would refer to the local variable.
-However, if the local variable declaration were removed, the second \ttt{x}
-would refer to the field instead.  This can be confusing to clients, since
-information collected about the program at one point in time may no longer be
-true after the tree's structure changes.
-\item \textit{It may be possible to mutate the AST so that it does not
-represent a legal program.}  It might even be possible to mutate the AST so
-that it is no longer a tree at all.  In Java, classes can contain other
-classes.  What happens if a client tries to make a class an inner class of
-itself?  (This could turn the tree into a cyclic graph.)  The API for modifying
-the AST's structure could try to prevent such things, but attempting to detect
-and prevent every possible mistake would be difficult and potentially
-expensive.
-\item \textit{Mutability can be problematic in multithreaded code.}
+    \item \textit{The API should describe the program's structure at a specific 
+        point in time, either before or after the a refactoring transformation.} 
+        It should never be inconsistent and provide some information that is
+        up-to-date with the latest transformation while providing some
+        informations that isn't.
 \end{itemize}
-
-Any candidate solution would have to satisfy the API requirements while avoiding
-the problems associated with mutable ASTs.
+\clearpage
+\begin{itemize}
+    \item \textit{The API should be robust in the presence of attempted refactorings
+        that produce invalid code.} The refactoring should never leave a 
+        client with code that doesn't compile. 
+    \item \textit{The API should be implemented in a way that allows easy analysis of
+        the syntactic effects of any given change.} This implementation should
+        be memory and time efficient, allowing clients to write refactorings
+        that meet the performance requirements of an interactive tool.
+    \item \textit{The API should be robust in the presence of "malicious" clients.}
+        its implementation should not allow clients to manipulate or 
+        invalidate internal data structures. This requires an understanding of
+        possible states of those data structures. In particular, an AST used
+        as an internal data structure might be exposed differently to clients
+        via the API. 
+\end{itemize}
 
@@
 
-\clearpage
 \section{Solution}
@@
 \subsection{Fulfilling the Requirements of Refactoring}
-\label{ss:reqs}
 
@@
 \subsection{Immutability}
+
+Since mutable ASTs are used by compilers, they seem like a reasonable mechanism for refactoring tools to use
+as well.  To remove a field declaration from a program, simply delete its
+\textit{FieldDeclaration} node from the AST (along with all its children).  To
+change the name of \ttt{FIVE}, change the contents of its
+\textit{SimpleNameNode} in the AST.  Of course, some mechanism is needed to
+prettyprint the modified AST, so these changes can be translated back into
+source code, but that is relatively straightforward to implement.
+
+Mutable ASTs have several benefits:
+\begin{itemize}
+    \item \textit{Mutable ASTs are simple to understand.} Performing 
+        refactorings on source code is conceptually identical to changing 
+        the AST itself.
+    \item \textit{Source code for specific refactorings can be very short.} 
+        The code simply has to rearrange the AST.
+
+\end{itemize}
+
+However, mutability introduces several issues:
+
+\FloatBarrier
+
+\begin{itemize}
+\item \textit{AST nodes contain source location information.}  After the
+structure of the AST is changed, what does the source location information in
+each node represent: the ``old'' source location (in the original code) or the
+``new'' location (after it has been modified)?  If it represents the old
+location, what should the source location be when new nodes are inserted into
+the tree?  If it represents the new location, then every time a change is made
+to the tree, source location information must be updated for every node in the
+tree, which may be expensive \dots or else a clever implementation (e.g.,
+computation from relative offsets) must be used instead.
+\item \textit{AST nodes may be removed from the tree, leaving them orphaned.}
+Suppose a client traverses the AST, storing pointers to some ``interesting''
+nodes in a collection.  Then, an ancestor of one of those nodes is removed from
+the AST.  This will leave the client in a difficult-to-debug situation where
+his collection contains pointers to nodes that no longer exist in the AST.
+\item \textit{Semantic information may be invalidated.}  Refactorings often
+need to determine, for a particular use of a variable, what declaration that
+use corresponds to.  For example, in Java, if a field and a local variable were
+both named \ttt{x}, then in the expression \ttt{this.x = x}, the first \ttt{x}
+would refer to the field, while the second would refer to the local variable.
+However, if the local variable declaration were removed, the second \ttt{x}
+would refer to the field instead.  This can be confusing to clients, since
+information collected about the program at one point in time may no longer be
+true after the tree's structure changes.
+\item \textit{It may be possible to mutate the AST so that it does not
+represent a legal program.}  It might even be possible to mutate the AST so
+that it is no longer a tree at all.  In Java, classes can contain other
+classes.  What happens if a client tries to make a class an inner class of
+itself?  (This could turn the tree into a cyclic graph.)  The API for modifying
+the AST's structure could try to prevent such things, but attempting to detect
+and prevent every possible mistake would be difficult and potentially
+expensive.
+\item \textit{Mutability can be problematic in multithreaded code.}
+\end{itemize}
 

### 2017-10-04 paper.tex
final changes

@@
         up-to-date with the latest transformation while providing some
-        informations that isn't.
+        information that isn't.
 \end{itemize}
@@
     \item \textit{The API should be robust in the presence of "malicious" clients.}
-        its implementation should not allow clients to manipulate or 
+        Its implementation should not allow clients to manipulate or 
         invalidate internal data structures. This requires an understanding of
@@
 %Maybe move to a different place?
-\subsection{Preprocessors}
-Preprocessors (such as the C preprocessor and M4), don't interfere with the ASTs
-used by compilers. In most cases, the preprocessing is performed first, and the
-output thereof is used to construct the AST. The lexer/parser requires no 
-knowledge of the fact that the text was preprocessed.
-
-Refactorings, however, are complicated by the C preprocessor. The grammar of 
-preprocessor directives doesn't take into account C syntax, and can't be easily
-represented in the C AST. This is the subject of Garrido's entire Ph.D.~thesis~%
-\cite{garrido05program}. The compiler's method of preprocessing before parsing
-can't be used, because the unpreprocessed text is what must be edited. The 
-refactoring too must be able to parse, analyze, and transform code \textit{with embedded preprocessor directives.}
-
-The topic of refactoring preprocessed code will not be treated in detail here,
-but suffice it to say that a compiler's AST will almost certainly be inadequate.
-For more details, see Garrido~\cite{garrido2013} and Overbey et al.~\cite{wrt2009}.
-
 \section{Solution}
@@
 unique to refactorings and require a pattern that specifically addresses them.
-
-\subsection{Source Mapping}
-
-The easiest way to map AST nodes to source locations is to store source
-location information in the AST nodes themselves.  Again, consider the JDT AST
-shown in Figure~\ref{fig:jdt-ast}, and note the numbers in square brackets.
-The highlighted node is displayed as ``VariableDeclarationFragment [35, 8]'':
-this AST node corresponds to the eight-character string beginning at offset 35
-(\ttt{FIVE = 5}).  Its child node corresponding to the identifier \ttt{FIVE} is
-attributed with the offset-length pair [35, 4], while the numeric literal
-\ttt{5}'s source location is [42, 1].
-
-Now, consider what happens when a user selects text in the text editor.  This
-can be represented as an offset-length pair.  When source location information
-is available for every AST node, it becomes fairly easy to determine what
-syntactic the text selection corresponds to: simply
-walk the tree, searching for a node whose source location aligns with the
-user's selection.
-
-Since AST nodes at deeper levels of the tree represent finer-grained syntactic
-constructs than their ancestor nodes, ASTs with source location information
-typically exhibit a nice property: the source location of a child node will
-always be contained within the source location of its parent.  This property
-can be used to expedite searches based on source location information.  For
-example, if a particular node corresponds to the source location [35, 8], and
-you are searching offset 63, then the desired node cannot possibly be a
-descendent of that node.
 
@@
 API, immutable trees are much more robust.
+
+\subsection{Source Mapping}
+
+The easiest way to map AST nodes to source locations is to store source
+location information in the AST nodes themselves.  Again, consider the JDT AST
+shown in Figure~\ref{fig:jdt-ast}, and note the numbers in square brackets.
+The highlighted node is displayed as ``VariableDeclarationFragment [35, 8]'':
+this AST node corresponds to the eight-character string beginning at offset 35
+(\ttt{FIVE = 5}).  Its child node corresponding to the identifier \ttt{FIVE} is
+attributed with the offset-length pair [35, 4], while the numeric literal
+\ttt{5}'s source location is [42, 1].
+
+Now, consider what happens when a user selects text in the text editor.  This
+can be represented as an offset-length pair.  When source location information
+is available for every AST node, it becomes fairly easy to determine what
+syntactic structure the text selection corresponds to: simply
+walk the tree, searching for a node whose source location aligns with the
+user's selection.
+
+Since AST nodes at deeper levels of the tree represent finer-grained syntactic
+constructs than their ancestor nodes, ASTs with source location information
+typically exhibit a nice property: the source location of a child node will
+always be contained within the source location of its parent.  This property
+can be used to expedite searches based on source location information.  For
+example, if a particular node corresponds to the source location [35, 8], and
+you are searching offset 63, then the desired node cannot possibly be a
+descendent of that node.
+
+This mapping, however, is complicated by the C preprocessor. The grammar of 
+preprocessor directives doesn't take into account C syntax, and can't be easily
+represented in the C AST. This is the subject of Garrido's entire Ph.D.~thesis~%
+\cite{garrido05program}. The topic of refactoring preprocessed code will not be treated in detail here,
+but suffice it to say that a compiler's AST will almost certainly be inadequate.
+For more details, see Garrido~\cite{garrido2013} and Overbey et al.~\cite{wrt2009}.
 
